---
title: 'Multiple Logistic Regression: Analyzing Cancer Risk'
author: "BioData Analysis Tutorial"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 5
)
```

---

# Introduction

**Logistic regression** is a statistical method used to model the relationship between one or more independent variables (predictors) and a **binary outcome** — an outcome with only two possible categories, usually coded as **0** and **1**.

Unlike linear regression (where the outcome is continuous), logistic regression predicts the **probability** that an event occurs — for example, whether a patient **does** or **does not** have cancer.

Because probabilities must lie between 0 and 1, logistic regression transforms the relationship using the **logit function**, which maps probabilities onto the entire real line:

$$\text{logit}(p) = \log\!\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k$$

The model **does not predict probability directly**, but predicts the **log odds**, which can then be transformed back:

$$\hat{p} = \frac{e^{\beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k}}{1 + e^{\beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k}}$$

---

## Study Design

Suppose you're a researcher studying factors associated with a positive cancer diagnosis. You collect data on **200 patients**. For each patient, you know:

| Variable | Type | Description |
|---|---|---|
| `Age` | Continuous | Patient age in years |
| `BMI` | Continuous | Body Mass Index |
| `SmokingStatus` | Categorical | Smoker / Non-Smoker |
| `FamilyHistory` | Categorical | Family history of cancer: Yes / No |
| `CancerStatus` | Binary outcome | 1 = diagnosed, 0 = not diagnosed |

---

# Step 1: Load Packages

```{r load-packages}

# Install missing packages and load libraries

packages <- c("tidyverse", "car", "pscl", "ResourceSelection", "pROC")

installed <- installed.packages()[, "Package"]
missing   <- packages[!packages %in% installed]

if (length(missing) > 0) {
  message("Installing missing packages: ", paste(missing, collapse = ", "))
  install.packages(missing)
} else {
  message("All packages already installed.")
}

library(tidyverse)
library(car)
library(pscl)
library(ResourceSelection)
library(pROC)

```


# Load the data
```{r data-summary}

data <- read_csv("framingham.csv")
# Quick overview
glimpse(data)

data=na.omit(data)

data$male <- as.factor(data$male)
data$currentSmoker <- as.factor(data$currentSmoker)
data$diabetes <- as.factor(data$diabetes)

# Outcome distribution
data %>%
  count(TenYearCHD) %>%
  mutate(Proportion = round(n / sum(n), 2))

```

---

# Step 2: Fit the Multiple Logistic Regression Model

```{r fit-model}
# Fit the logistic regression model
model <- glm(TenYearCHD ~ age + BMI + male + currentSmoker + diabetes + totChol + glucose,
             data   = data,
             family = binomial)

# View the model summary
summary(model)
```

---

# Step 3: Interpret the Results

## Odds Ratios

Coefficients are on the **log-odds scale**. Exponentiating converts them to **odds ratios (OR)** — the most intuitive unit of interpretation.

```{r odds-ratios}
# Compute odds ratios with 95% confidence intervals
OR_table <- exp(cbind(OR = coef(model), confint(model)))
round(OR_table, 3)
```

**Interpretation of each predictor:**



## Coefficient Plot

```{r coef-plot}
# Build a tidy OR data frame for plotting
or_df <- data.frame(
  Term     = rownames(OR_table),
  OR       = OR_table[, "OR"],
  Lower    = OR_table[, "2.5 %"],
  Upper    = OR_table[, "97.5 %"]
) %>%
  filter(Term != "(Intercept)")

ggplot(or_df, aes(x = OR, y = reorder(Term, OR))) +
  geom_point(size = 4, color = "tomato") +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper), height = 0.25, color = "tomato") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  labs(title = "Odds Ratios with 95% Confidence Intervals",
       x = "Odds Ratio (OR)", y = NULL) +
  theme_minimal(base_size = 13)
```

---

# Step 4: Visualize Model Performance (ROC & AUC)

The **Receiver Operating Characteristic (ROC) curve** shows how well the model discriminates between patients with and without cancer across all possible thresholds.

```{r roc-curve}
# Get predicted probabilities from the model
probabilities <- predict(model, type = "response")

# Actual outcomes
actual <- data$TenYearCHD

# Create ROC object
roc_obj <- roc(actual, probabilities, quiet = TRUE)

# Plot ROC curve
plot(roc_obj,
     main = "ROC Curve for Cancer Prediction",
     col  = "#2C7BB6",
     lwd  = 3)
legend("bottomright",
       legend = paste0("AUC = ", round(auc(roc_obj), 3)),
       col = "#2C7BB6", lwd = 3, bty = "n")

# Print AUC
cat("Area under the curve (AUC):", round(auc(roc_obj), 3), "\n")
```

**AUC Interpretation guide:**

| AUC | Discrimination |
|---|---|
| 0.6 | No discrimination (random) |
| 0.7 – 0.8 | Acceptable |
| 0.8 – 0.9 | Good |
| > 0.9 | Excellent |

An AUC of **~0.799** indicates the model can reliably rank patients by cancer risk **79.9% of the time**.

---

# Step 6: Confusion Matrix & Classification Metrics

Choose a **threshold** (default = 0.5) to convert predicted probabilities into class labels.

```{r confusion-matrix}
# Classify patients using 0.5 threshold
predicted_class <- ifelse(probabilities > 0.5, 1, 0)

# Confusion matrix
cm <- table(Predicted = predicted_class, Actual = actual)
print(cm)
```

```{r classification-metrics}
# Accuracy
accuracy    <- mean(predicted_class == actual)

# Sensitivity (True Positive Rate — how well we catch cancer cases)
sensitivity <- sum(predicted_class == 1 & actual == 1) / sum(actual == 1)

# Specificity (True Negative Rate — how well we rule out cancer)
specificity <- sum(predicted_class == 0 & actual == 0) / sum(actual == 0)

cat(sprintf("Accuracy    : %.2f\n", accuracy))
cat(sprintf("Sensitivity : %.2f\n", sensitivity))
cat(sprintf("Specificity : %.2f\n", specificity))
```

**Interpretation:**

- **True Negatives**: Patients correctly classified as no cancer.
- **True Positives**: Patients correctly classified as having cancer.
- **False Negatives**: Cancer cases the model missed *(clinically costly)*.
- **False Positives**: Non-cancer patients flagged as at risk.


---

# Step 6: Check Assumptions (Diagnostics)

## 7.1 Linearity in the Logit (Box-Tidwell Test)

Check whether continuous predictors have a linear relationship with the log-odds by adding interaction terms $x \cdot \log(x)$.

```{r box-tidwell}
# Add log-transformed terms
data$logage <- log(data$age)
data$logBMI <- log(data$BMI)
data$logglucose <- log(data$glucose)

# Fit model with Box-Tidwell interaction terms
model_bt <- glm(TenYearCHD ~ age + BMI + male + currentSmoker + diabetes + totChol + glucose +
                  I(age * logage) + I(BMI * logBMI) + I(glucose*logglucose),
                data   = data,
                family = binomial)

summary(model_bt)
```

> If `I(Age * logAge)` and `I(BMI * logBMI)` are **not significant**, the linearity assumption holds. ✔

## 6.2 Multicollinearity (VIF)

```{r vif}
# Variance Inflation Factors
vif(model)
```

> VIF values **< 5** indicate acceptable multicollinearity. Values near 1–2 are ideal.

## 6.3 Influential Points (Cook's Distance)

```{r cooks-distance}
cooksd <- cooks.distance(model)

plot(cooksd,
     pch  = 19,
     cex  = 1.2,
     col  = ifelse(cooksd > 4 / nrow(data), "tomato", "steelblue"),
     main = "Cook's Distance — Influential Observations",
     ylab = "Cook's Distance")
abline(h = 4 / nrow(data), col = "red", lty = 2)

# Identify influential points
influential <- which(cooksd > 4 / nrow(data))
cat("Number of influential points:", length(influential), "\n")
if (length(influential) > 0) cat("Row indices:", influential, "\n")
```

---

# Step 7: Goodness-of-Fit

## 7.1 McFadden's Pseudo R²

```{r pseudo-r2}
pR2(model)
```

> McFadden's R² values of **0.2–0.4** suggest a reasonably well-fitting model.

## 7.2 Likelihood Ratio Test

Compare the fitted model to a **null model** (intercept only) to confirm that the predictors improve fit.

```{r lrt}
# Null (intercept-only) model
model_null <- glm(TenYearCHD ~ 1, data = data, family = binomial)

# Likelihood Ratio Test
anova(model_null, model, test = "Chisq")
```

> A **significant p-value** means the predictors collectively improve the model beyond the null. ✔

## 7.3 Hosmer-Lemeshow Test

```{r hosmer-lemeshow}
# Hosmer-Lemeshow goodness-of-fit test (g = 10 groups)
hoslem.test(data$TenYearCHD, fitted(model), g = 10)
```

> A **non-significant p-value (> 0.05)** indicates good calibration — predicted probabilities match observed event rates.



---

# Part II — Poisson Regression


## Model Building


### Model 3 — Adjusted 

```{r epil-adjusted}
pois_model <- glm(cigsPerDay ~ male + age + BMI,
           data   = data,
           family = poisson)
summary(pois_model)
```

```{r}
IRR_table <- exp(cbind(
  IRR      = coef(pois_model),
  Lower95  = confint(pois_model)[, 1],
  Upper95  = confint(pois_model)[, 2]
))

cat("\nIRR with 95% CI:\n")
print(round(IRR_table, 3))
```

Interpretation: 


#Plot the results 

```{r epil-irr-forest}
cf_ep2 <- data.frame(
  Term     = names(coef(pois_model))[-1],
  IRR      = exp(coef(pois_model))[-1],
  CI_lower = exp(confint.default(pois_model))[-1, 1],
  CI_upper = exp(confint.default(pois_model))[-1, 2],
  p_value  = coef(summary(pois_model))[-1, 4]
)

ggplot(cf_ep2, aes(x = IRR, y = reorder(Term, IRR))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper),
                 height = 0.25, color = "#2c7fb8", linewidth = 0.9) +
  geom_point(aes(color = p_value < 0.05), size = 3.5) +
  scale_color_manual(
    values = c("TRUE" = "#d7301f", "FALSE" = "#2c7fb8"),
    labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05"), name = NULL
  ) +
  scale_x_log10() +
  labs(
    title    = "Incidence Rate Ratios: Epileptic Seizures (Adjusted Poisson Model)",
    subtitle = "IRR < 1 = fewer seizures; IRR > 1 = more seizures",
    x = "Incidence Rate Ratio (log scale)", y = NULL
  ) +
  theme(legend.position = "top")
```



## Diagnostics

### 1. Overdispersion Check

```{r epil-overdispersion}
phi <- deviance(pois_model) / df.residual(pois_model)
p_formal <- pchisq(deviance(pois_model), df.residual(pois_model), lower.tail = FALSE)

cat("── Overdispersion Diagnostic ──────────────────────────────────\n")
cat(sprintf("Residual deviance:  %.2f\n", deviance(pois_model)))
cat(sprintf("Residual df:        %d\n",   df.residual(pois_model)))
cat(sprintf("Dispersion φ = Dev/df: %.3f\n\n", phi))
cat(sprintf("Formal chi-square test p-value: %.8f\n\n", p_formal))
cat(sprintf("Interpretation: φ = %.2f >> 1 → strong overdispersion detected.\n", phi))
cat("Standard Poisson SEs are underestimated → inference is anti-conservative.\n")
cat("Remedies: Quasi-Poisson (corrects SE) or Negative Binomial (re-specifies variance).\n")
```

### 2. QQ Plot of Deviance Residuals

```{r epil-qq}
qqnorm(residuals(pois_model, type = "deviance"),
       main  = "QQ Plot of Deviance Residuals (Poisson Model)",
       col   = "#2c7fb8", pch = 16, cex = 0.7)
qqline(residuals(pois_model, type = "deviance"),
       col = "#d7301f", lwd = 1.5)
mtext("Heavy tails confirm overdispersion — Poisson assumption is violated",
      side = 3, cex = 0.8, col = "gray40")
```

