---
title: "Introduction to R & Data Management"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  html_notebook:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
subtitle: A Clinician's First Steps — Illustrated with Patient Data
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width  = 7.5,
  fig.height = 4.5,
  fig.align  = "center"
)
```


------------------------------------------------------------------------

**Learning objectives**

1.  Navigate the R/RStudio environment and understand the core R syntax
2.  Work with R's fundamental data types and structures
3.  Build and manipulate data frames that mirror a patient registry
4.  Import, inspect, and clean messy clinical data
5.  Reshape longitudinal patient data between wide and long formats
6.  Summarise and export a clean analytic dataset

------------------------------------------------------------------------

# The R Environment

## Why R in Healthcare?

R is free, reproducible, and used across clinical research,
epidemiology, public health, and regulatory submissions (FDA accepts
R-based analyses). The same script that cleaned your data documents
exactly what was done — essential for audit trails.

## The RStudio Interface

| Pane | Purpose |
|------------------------------------|------------------------------------|
| **Source** (top-left) | Write and save `.R` / `.Rmd` scripts |
| **Console** (bottom-left) | Run code interactively; see output |
| **Environment** (top-right) | See all objects currently in memory |
| **Files / Plots / Help** (bottom-right) | Browse files, view plots, read docs |

## Your First Commands

```{r first-commands}
# Lines starting with # are comments — R ignores them
# Comments are your notes inside the code

2 + 2          # arithmetic
sqrt(144)      # square root
log(10)        # natural log
log10(10)      # base-10 log
abs(-7.4)      # absolute value
```

## Assigning Objects

```{r assignment}
# <- is the assignment operator  (keyboard shortcut: Alt + -)
patient_age  <- 54
patient_name <- "Maria Silva"
is_diabetic  <- FALSE

# Print by typing the name
patient_age
patient_name
is_diabetic

# Or use cat() / print() for formatted output
cat("Patient:", patient_name, "| Age:", patient_age, "\n")
```

::: tip
Avoid spaces and special characters. Names are case-sensitive:
<code>BMI</code> ≠ <code>bmi</code>.
:::

------------------------------------------------------------------------

# Data Types

Choosing the wrong type of variable is one of
the most common sources of errors in data analysis.

## The Four Core Types

```{r types-demo}
# numeric — continuous measurements
systolic_bp   <- 128.5
body_temp_c   <- 37.2
creatinine    <- 1.04

# character — text / identifiers
patient_id    <- "PT-00847"
diagnosis     <- "Type 2 Diabetes Mellitus"
icd10_code    <- "E11.9"

# logical — binary TRUE / FALSE flags
on_insulin    <- FALSE
icu_admission <- TRUE
allergy_flag  <- NA          # NA = missing / not recorded

# integer — whole numbers (add L suffix)
ward_number   <- 4L
hospital_days <- 7L

# Check the type
class(systolic_bp)
class(patient_id)
class(on_insulin)
class(ward_number)
```

## Factors — Categorical Clinical Variables

Factors represent variables with a fixed set of levels (sex, blood
group, disease severity). They are essential for correct statistical
modelling.

```{r factors}
# Create a factor
blood_group <- factor("A",
                      levels = c("A", "B", "AB", "O"))

severity <- factor("Moderate",
                   levels = c("Mild", "Moderate", "Severe"),
                   ordered = TRUE)   # ordered = TRUE for ordinal variables

# Inspect
levels(blood_group)
levels(severity)
nlevels(severity)

```

## Type Coercion — Converting Between Types

```{r coercion}
# Character to numeric (e.g., data imported from CSV as text)
sbp_text   <- "132"
sbp_number <- as.numeric(sbp_text)
sbp_number + 10

# Numeric to factor (e.g., coding groups)
diabetes_code <- c(0, 1, 0, 1, 1)
diabetes_fct  <- factor(diabetes_code,
                         levels = c(0, 1),
                         labels = c("No", "Yes"))
diabetes_fct

# What happens when conversion fails? Produces NA with a warning
```

------------------------------------------------------------------------

# Vectors — Working with Patient Measurements

A **vector** is a sequence of values of the same type — the building
block of all R data structures.

## Creating Vectors

```{r vectors-create}
# Systolic blood pressure readings for 8 patients
sbp <- c(118, 135, 142, 128, 156, 101, 148, 122)

# Patient IDs
ids <- c("PT-001", "PT-002", "PT-003", "PT-004",
         "PT-005", "PT-006", "PT-007", "PT-008")

# Diabetes status
diabetic <- c(FALSE, TRUE, TRUE, FALSE, TRUE, FALSE, TRUE, FALSE)

# Sequences — useful for generating IDs or visit numbers
visit_num <- 1:6                      # 1 2 3 4 5 6
age_grid  <- seq(20, 80, by = 10)    # 20 30 40 50 60 70 80

```

## Vectorised Operations — No Loops Needed

```{r vectors-ops}
# Convert SBP from mmHg to kPa (1 mmHg ≈ 0.1333 kPa)
sbp_kPa <- sbp * 0.1333
sbp_kPa

# Classify hypertension (SBP ≥ 140 mmHg)
is_hypertensive <- sbp >= 140
is_hypertensive

# How many patients are hypertensive?
sum(is_hypertensive)           # TRUE counts as 1
mean(is_hypertensive) * 100    # percentage
```

## Subsetting Vectors — Selecting Patients

```{r vectors-subset}
# By position (1-indexed in R)
sbp[1]           # first patient
sbp[c(1, 3, 5)]  # patients 1, 3, 5

# By condition (logical subsetting)
sbp[is_hypertensive]            # SBP values of hypertensive patients
ids[is_hypertensive]            # IDs of hypertensive patients
sbp[sbp > 130 & sbp < 150]     # between 130 and 150

# By name (named vectors)
names(sbp) <- ids
sbp["PT-003"]
```

## Descriptive Functions

```{r vectors-stats}
length(sbp)          # sample size
sum(sbp)
mean(sbp)            # arithmetic mean
median(sbp)
sd(sbp)              # standard deviation
var(sbp)             # variance
range(sbp)           # min and max
quantile(sbp)        # quartiles
IQR(sbp)             # interquartile range
summary(sbp)         # all in one
```

------------------------------------------------------------------------

# Data Frames — The Patient Registry 

A **data frame** is a table where each row is a patient (observation)
and each column is a variable. This mirrors a clinical database exactly.

## Building a Patient Registry from Scratch

```{r df-create}
patients <- data.frame(
  patient_id  = c("PT-001","PT-002","PT-003","PT-004","PT-005",
                  "PT-006","PT-007","PT-008","PT-009","PT-010"),
  age         = c(54, 67, 45, 72, 38, 61, 55, 49, 80, 33),
  sex         = factor(c("F","M","F","M","F","M","F","M","M","F"),
                       levels = c("F","M")),
  systolic_bp = c(118, 142, 135, 158, 112, 148, 130, 122, 165, 105),
  diastolic_bp= c( 76,  92,  88,  98,  70,  95,  84,  78, 102,  68),
  bmi         = c(24.1, 29.8, 32.4, 27.0, 21.5, 31.2, 26.8, 23.9, 28.6, 20.1),
  diabetes    = factor(c("No","Yes","Yes","No","No","Yes","No","No","Yes","No"),
                       levels = c("No","Yes")),
  on_medication = c(FALSE, TRUE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE, TRUE, FALSE),
  hba1c       = c(5.4, 8.1, 7.6, 5.9, 5.2, 8.8, 5.7, 5.5, 9.2, 4.9),
  stringsAsFactors = FALSE
)

patients

```

## Inspecting the Registry

```{r df-inspect}
nrow(patients)      # number of patients
ncol(patients)      # number of variables
dim(patients)       # both at once

str(patients)       # structure — types + preview (always run this after importing)
head(patients, 4)   # first 4 rows
tail(patients, 3)   # last 3 rows
summary(patients)   # statistical overview per column
```

## Accessing Columns

```{r df-cols}
patients$age                  # dollar sign — returns a vector
patients$sex
patients[["systolic_bp"]]     # double-bracket — identical result
patients[ , "bmi"]            # [row, col] — blank row = all rows
patients[ , c("patient_id", "age", "diabetes")]   # multiple columns
```

## Accessing Rows — Querying Patients

```{r df-rows}
patients[3, ]                                 # row 3, all columns
patients[patients$diabetes == "Yes", ]        # all diabetic patients
patients[patients$age > 60, ]                 # patients over 60
patients[patients$bmi > 30 & patients$sex == "M", ]  # obese male patients
```

## Adding & Modifying Columns

```{r df-mutate}
# Derived variable: pulse pressure
patients$pulse_pressure <- patients$systolic_bp - patients$diastolic_bp

# Classify BP stage (JNC 8 thresholds)
patients$bp_stage <- cut(
  patients$systolic_bp,
  breaks = c(0, 119, 129, 139, Inf),
  labels = c("Normal", "Elevated", "Stage 1 HTN", "Stage 2 HTN"),
  right  = TRUE
)

# Recode HbA1c control: <7% = controlled
patients$hba1c_controlled <- ifelse(patients$hba1c < 7, "Controlled", "Uncontrolled")

patients[, c("patient_id","systolic_bp","pulse_pressure","bp_stage","hba1c","hba1c_controlled")]
```


------------------------------------------------------------------------

# Packages & the Tidyverse {.badge}

```{r pkg-install, eval=FALSE}
# Run once to install, then comment using #
#install.packages(c("tidyverse", "readxl", "janitor", "skimr", "writexl", "lubridate"))
```


```{r pkg-load}
#Load the packages
library(tidyverse)    # dplyr + tidyr + ggplot2 + readr + more
library(janitor)      # clean_names(), tabyl()
library(skimr)        # skim() — rich summary statistics
library(lubridate)    # date handling
library(readxl)       # read Excel files
library(writexl)      # write Excel files
```

| Package     | Role in data management                               |
|-------------|-------------------------------------------------------|
| `dplyr`     | Filter, select, mutate, summarise patient records     |
| `tidyr`     | Reshape wide ↔ long (e.g., repeated measurements)     |
| `readr`     | Read CSV files quickly with smart type detection      |
| `readxl`    | Read `.xlsx` / `.xls` files                           |
| `janitor`   | Clean messy column names, frequency tables            |
| `lubridate` | Parse and calculate with dates (admission, DOB, etc.) |
| `skimr`     | Rich one-call data summary                            |

------------------------------------------------------------------------

# Importing Clinical Data

## Reading CSV Files

```{r import-csv, eval=FALSE}

stwd("C:/Users/ibrah/OneDrive/Desktop/Physalia Course")

#Save the dataframe as csv file in working directory
write.csv(patients, "patient_registry.csv")

# From a local file
patient_data <- read_csv("patient_registry.csv") #importing dataset

# Common arguments
patient_data <- read_csv(
  "data/patient_registry.csv",
  na      = c("", "NA", "N/A", "missing", "9999"),  # what counts as missing
  skip    = 2,         # skip 2 header rows (common in exported EMR files)
  n_max   = 5000       # read first 5000 rows only
)


```

## Simulating a "Messy Import" — What Real Data Looks Like

```{r import-messy}
# Simulate what you'd get from a poorly formatted hospital export
messy_raw <- data.frame(
  `Patient ID`     = c("pt001", "PT002", "pt003", "PT004", "pt005",
                        "pt006", "PT007", "pt008", "pt009", "pt010"),
  `Age (years)`    = c("23", "45", "67", "34", "56", "29", "72", "41", "38", "N/A"),
  `SEX`            = c("Male","female","MALE","Female","male",
                        "FEMALE","Male","female","Male","Female"),
  `SBP (mmHg)`     = c("120", "138", "155", "112", "140",
                        "125", "162", "130", "118", "145"),
  `Admission Date` = c("01/03/2024","15-03-2024","2024-03-20","2024/03/22","03/04/2024",
                        "10-04-2024","2024-04-15","25/04/2024","30-04-2024","2024-05-01"),
  `Outcome`        = c("discharged","Discharged","DISCHARGED","deceased",
                        "discharged","transferred","Discharged","deceased",
                        "Discharged","TRANSFERRED"),
  check.names      = FALSE,
  stringsAsFactors = FALSE
)

messy_raw
```

Notice the problems: inconsistent capitalisation, mixed date formats,
ages stored as characters, column names with spaces and units, "N/A" not
recognised as missing.

------------------------------------------------------------------------

# Data Cleaning

## Step 1 — Fix Column Names

```{r clean-names}
# janitor::clean_names() converts all names to snake_case
clean1 <- messy_raw |>
  clean_names()

names(clean1)
```

## Step 2 — Fix Data Types

```{r clean-types}
clean2 <- clean1 |>
  rename(
    patient_id = patient_id,
    age        = age_years,
    sbp        = sbp_mm_hg,
    adm_date   = admission_date
  ) |>
  mutate(
    age = as.numeric(age),                    # "N/A" becomes NA automatically
    sbp = as.numeric(sbp)
  )

str(clean2)
```

## Step 3 — Standardise Categorical Values

```{r clean-categories}
clean3 <- clean2 |>
  mutate(
    # Standardise sex to Title Case, then factor
    sex     = str_to_title(str_trim(sex)),
    sex     = factor(sex, levels = c("Female", "Male")),

    # Standardise outcome
    outcome = str_to_lower(str_trim(outcome)),
    outcome = factor(outcome,
                     levels = c("discharged", "deceased", "transferred"))
  )

count(clean3, sex)
count(clean3, outcome)
```

## Step 4 — Parse Dates

```{r clean-dates}
clean4 <- clean3 |>
  mutate(
    # lubridate::parse_date_time handles multiple mixed formats in one call
    adm_date = parse_date_time(
      adm_date,
      orders = c("dmy", "dmY", "Ymd", "Y/m/d")
    ) |> as_date()
  )

clean4 |> select(patient_id, adm_date)

# Date arithmetic
clean4 <- clean4 |>
  mutate(
    days_since_adm = as.numeric(today() - adm_date),
    adm_month      = month(adm_date, label = TRUE, abbr = FALSE),
    adm_year       = year(adm_date)
  )

clean4 |> select(patient_id, adm_date, days_since_adm, adm_month)
```

## Step 5 — Handle Missing Values

```{r clean-missing}
# How many missing per column?
colSums(is.na(clean4))

# Visualise missingness pattern
clean4 |>
  summarise(across(everything(), ~mean(is.na(.)) * 100)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
  arrange(desc(pct_missing))
```

```{r clean-impute}
# Option A — Drop rows with missing key variables
clean4 |> drop_na(age, sbp)

# Option B — Replace NA with median (simple imputation for teaching)
clean5 <- clean4 |>
  mutate(
    age = if_else(is.na(age), median(age, na.rm = TRUE), age),
    sbp = if_else(is.na(sbp), median(sbp, na.rm = TRUE), sbp)
  )

colSums(is.na(clean5))
```

## Step 6 — Remove Duplicates

```{r clean-dupes}
# Check for duplicate patient IDs
clean5 |>
  count(patient_id) |>
  filter(n > 1)

# If duplicates exist: keep the first occurrence
clean6 <- clean5 |>
  distinct(patient_id, .keep_all = TRUE)

nrow(clean5)
nrow(clean6)
```

------------------------------------------------------------------------

# Data Manipulation with dplyr {.badge}

We switch back to the clean `patients` data frame (10 patients with
known clinical variables) for these examples.

## The Pipe Operator `|>`

The pipe passes the left-hand side as the first argument to the next
function:

```{r pipe-demo}
# Without pipe — hard to read
round(mean(patients$systolic_bp), 1)

# With pipe — reads left to right like a protocol
patients$systolic_bp |> mean() |> round(1)
```

## filter() — Select Rows (Patients)

```{r dplyr-filter}
# Diabetic patients only
patients |> filter(diabetes == "Yes")

# Hypertensive women over 50
patients |> filter(systolic_bp >= 140, sex == "F", age > 50)

# Multiple allowed values (like SQL IN)
patients |> filter(bp_stage %in% c("Stage 1 HTN", "Stage 2 HTN"))
```

## select() — Choose Variables

```{r dplyr-select}
# Keep only the variables needed for a BP analysis
patients |> select(patient_id, age, sex, systolic_bp, diastolic_bp, bp_stage)

# Drop variables with -
patients |> select(-on_medication, -hba1c_controlled)

# Reorder: bring key ID columns first
patients |> select(patient_id, age, sex, everything())

# Select by type
patients |> select(where(is.numeric))
```

## mutate() — Create & Transform Variables

```{r dplyr-mutate}
patients <- patients |>
  mutate(
    # BMI classification (WHO thresholds)
    bmi_class = case_when(
      bmi < 18.5             ~ "Underweight",
      bmi >= 18.5 & bmi < 25 ~ "Normal",
      bmi >= 25   & bmi < 30 ~ "Overweight",
      bmi >= 30              ~ "Obese"
    ) |> factor(levels = c("Underweight","Normal","Overweight","Obese"),
                ordered = TRUE),

    # Age group — common in clinical reporting
    age_group = cut(age,
                    breaks = c(0, 40, 60, Inf),
                    labels = c("<40", "40–60", ">60"),
                    right  = FALSE),

    # Flag high-risk: diabetic AND hypertensive
    high_risk = diabetes == "Yes" & systolic_bp >= 140
  )

patients |> select(patient_id, age, age_group, bmi, bmi_class, diabetes,
                   systolic_bp, high_risk)
```

## arrange() — Sort Records

```{r dplyr-arrange}
# By SBP descending — find most hypertensive patients
patients |>
  arrange(desc(systolic_bp)) |>
  select(patient_id, age, sex, systolic_bp, bp_stage)
```

## summarise() + group_by() — Clinical Aggregation

```{r dplyr-summarise}
# Overall summary statistics
patients |>
  summarise(
    n          = n(),
    mean_sbp   = round(mean(systolic_bp), 1),
    sd_sbp     = round(sd(systolic_bp), 1),
    pct_diab   = round(mean(diabetes == "Yes") * 100, 1),
    mean_bmi   = round(mean(bmi), 1)
  )

# Stratified by diabetes status
patients |>
  group_by(diabetes) |>
  summarise(
    n        = n(),
    mean_sbp = round(mean(systolic_bp), 1),
    mean_bmi = round(mean(bmi), 1),
    mean_hba1c = round(mean(hba1c), 1)
  )

# Three-way breakdown: sex × diabetes
patients |>
  group_by(sex, diabetes) |>
  summarise(n = n(), mean_sbp = round(mean(systolic_bp), 1),
            .groups = "drop")
```

## Full Cleaning Pipeline — Chaining Steps

```{r dplyr-pipeline}
# Build an analytic sub-dataset for diabetic patients in one readable pipeline
diabetic_summary <- patients |>
  filter(diabetes == "Yes") |>
  select(patient_id, age, sex, bmi, bmi_class, hba1c,
         systolic_bp, bp_stage, on_medication) |>
  mutate(
    bp_controlled  = systolic_bp < 130,
    hba1c_target   = hba1c < 7.0
  ) |>
  arrange(desc(hba1c))

diabetic_summary
```

------------------------------------------------------------------------

# Reshaping Data with tidyr {.badge}

Clinical data often arrives in **wide** format (one row per patient, one
column per time-point) but analyses require **long** format (one row per
patient-visit).

## Wide → Long: Repeated BP Measurements

```{r reshape-wide}
# Wide format: SBP measured at 4 visits
bp_wide <- tibble(
  patient_id = c("PT-001","PT-002","PT-003","PT-004","PT-005"),
  sbp_wk0    = c(158, 145, 172, 138, 165),
  sbp_wk4    = c(148, 138, 160, 132, 154),
  sbp_wk8    = c(140, 130, 152, 128, 145),
  sbp_wk12   = c(132, 125, 144, 122, 138)
)

bp_wide
```

```{r pivot-longer}
# Long format: one row per patient per visit
bp_long <- bp_wide |>
  pivot_longer(
    cols         = starts_with("sbp_wk"),
    names_to     = "visit",
    names_prefix = "sbp_",         # strip the prefix to keep "wk0", "wk4"…
    values_to    = "sbp"
  ) |>
  mutate(
    week = as.integer(str_remove(visit, "wk"))  # extract numeric week
  )

bp_long
```

```{r pivot-longer-plot}
# Long format enables ggplot2 to draw a trajectory per patient
ggplot(bp_long, aes(x = week, y = sbp,
                    colour = patient_id, group = patient_id)) +
  geom_line(linewidth = .9) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 130, linetype = "dashed",
             colour = "grey40", linewidth = .7) +
  annotate("text", x = 10.5, y = 131.5,
           label = "Target < 130 mmHg", size = 3.2, colour = "grey40") +
  scale_x_continuous(breaks = c(0, 4, 8, 12)) +
  labs(title   = "Systolic BP Trajectories Over 12 Weeks",
       x       = "Week",
       y       = "Systolic BP (mmHg)",
       colour  = "Patient") +
  theme_minimal(base_size = 13)
```

## Long → Wide: Spreading Back

```{r pivot-wider}
# Useful when creating summary tables for reports
bp_long |>
  select(patient_id, visit, sbp) |>
  pivot_wider(names_from  = visit,
              values_from = sbp,
              names_prefix = "sbp_")
```

## separate() & unite() — Splitting / Combining Columns

```{r separate-unite}
# A column mixing ward and bed number
location <- tibble(location_code = c("ICU-12","Ward3-07","ED-03","Ward2-14"))

# Split into two columns
location |>
  separate(location_code, into = c("ward", "bed"), sep = "-")

# Join two columns into one (e.g., first + last name)
names_df <- tibble(first = c("Maria","João","Ana"), last = c("Silva","Costa","Pereira"))
names_df |>
  unite("full_name", first, last, sep = " ")
```

------------------------------------------------------------------------

# Joining Datasets 

Clinical data often lives in separate tables (demographics, labs,
prescriptions) that must be linked by a patient ID.

```{r joins-setup}
# Table 1: demographics
demographics <- tibble(
  patient_id = c("PT-001","PT-002","PT-003","PT-004","PT-005"),
  age        = c(54, 67, 45, 72, 38),
  sex        = c("F","M","F","M","F")
)

# Table 2: most recent lab results (not all patients have labs)
labs <- tibble(
  patient_id  = c("PT-001","PT-002","PT-004","PT-006"),   # PT-003/005 missing, PT-006 new
  hba1c       = c(7.1, 8.4, 6.2, 9.1),
  creatinine  = c(0.9, 1.2, 1.0, 1.8)
)
```

```{r joins-demo}
# left_join: keep ALL patients from demographics, add labs where available
left_join(demographics, labs, by = "patient_id")

# inner_join: keep only patients present in BOTH tables
inner_join(demographics, labs, by = "patient_id")

# anti_join: patients in demographics but NOT in labs (missing labs)
anti_join(demographics, labs, by = "patient_id")
```

| Join type | Keeps |
|------------------------------------|------------------------------------|
| `left_join` | All rows from left table; matched data from right (fill NA if no match) |
| `right_join` | All rows from right table |
| `inner_join` | Only rows with a match in both tables |
| `full_join` | All rows from both tables |
| `anti_join` | Rows in left with NO match in right (find missing records) |

------------------------------------------------------------------------

# Saving Your Work {.badge}

```{r save-csv, eval=FALSE}
# Export the clean analytic dataset as CSV
write_csv(patients, "clean_patient_registry.csv")

# Export as Excel (requires writexl package)
# install.packages("writexl")
writexl::write_xlsx(patients, "clean_patient_registry.xlsx")
```

```{r save-rds, eval=FALSE}
# Save as an R object (preserves factor levels, data types exactly)
saveRDS(patients, "clean_patient_registry.rds")

# Reload later
patients <- readRDS("clean_patient_registry.rds")
```

```{r save-plot, eval=FALSE}
# Save the last ggplot
ggsave("bp_trajectories.png", width = 8, height = 5, dpi = 150)
```

------------------------------------------------------------------------

# Exercises {.tabset}

## Exercise 1 — Vector Operations

**You have HbA1c values for 10 patients:**\
`c(5.4, 8.1, 7.6, 5.9, 5.2, 8.8, 5.7, 5.5, 9.2, 4.9)`

1.  Store them in a vector called `hba1c`
2.  How many patients have HbA1c ≥ 7.0 (diabetic range)?
3.  What is the mean HbA1c of those patients?

### Answer

```{r ex1}
#Your code here

```

## Exercise 2 — Data Frame + dplyr

**Using the `patients` data frame:**

1.  Filter to patients who are obese (`bmi_class == "Obese"`) and on
    medication
2.  Select only `patient_id`, `age`, `sex`, `bmi`, `hba1c`,
    `systolic_bp`
3.  Arrange by `hba1c` descending
4.  Add a column `bp_risk` that is `TRUE` if `systolic_bp ≥ 140`

### Answer

```{r ex2}
#your code here


```

## Exercise 3 — Pivot + Visualise

**The table below shows eGFR (kidney function) measured at three
time-points.**\
Pivot it to long format, then plot eGFR trajectories over time.

```{r ex3-setup}
egfr_wide <- tibble(
  patient_id  = paste0("PT-00", 1:6),
  egfr_month0 = c(85, 62, 45, 72, 90, 55),
  egfr_month6 = c(80, 58, 42, 70, 88, 51),
  egfr_month12= c(74, 53, 38, 68, 85, 47)
)
egfr_wide
```

### Answer

```{r ex3}
#Your code here


```

## Exercise 4 — Joining Tables

**Join the `demographics` and `labs` tables (created in the Joining
section).**

1.  Perform a `left_join` and count how many patients are missing lab
    values
2.  Create a flag column `labs_available` that is `TRUE` when `hba1c` is
    not NA

### Answer

```{r ex4}
#Your code here


```

------------------------------------------------------------------------

# Summary — What You Learned

| Topic | Key functions |
|------------------------------------|------------------------------------|
| **R syntax & objects** | `<-`, `class()`, `c()`, `factor()`, `as.numeric()` |
| **Vector operations** | subsetting with `[]`, `sum()`, `mean()`, `summary()` |
| **Data frames** | `data.frame()`, `$`, `str()`, `head()`, `nrow()` |
| **Importing data** | `read_csv()`, `read_excel()`, `clean_names()` |
| **Data cleaning** | `str_to_lower()`, `parse_date_time()`, `drop_na()`, `distinct()` |
| **dplyr verbs** | `filter()`, `select()`, `mutate()`, `arrange()`, `summarise()`, `group_by()` |
| **Reshaping** | `pivot_longer()`, `pivot_wider()`, `separate()`, `unite()` |
| **Joining** | `left_join()`, `inner_join()`, `anti_join()` |
| **Exporting** | `write_csv()`, `saveRDS()`, `ggsave()` |

::: info
<strong>📚 Recommended next steps</strong><br> • <strong>R for Data
Science (free):</strong> <https://r4ds.hadley.nz> — chapters on data
transformation and tidy data<br> • <strong>CRAN Task View — Clinical
Trials:</strong> <https://cran.r-project.org/view=ClinicalTrials><br> •
<strong>dplyr cheatsheet:</strong>
<https://posit.co/resources/cheatsheets/><br> • <strong>Biostatistics
with R (this series, next session):</strong> t-tests, chi-square,
regression on the NHANES dataset
:::

```{r session-info, echo=FALSE}
sessionInfo()
```
