---
title: "Linear Models — ANOVA and Regression"
subtitle: "Fundamentals of Biostatistics — Hands-on Session 7"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

---

# Overview

This session introduces linear models as a unified framework that covers both
ANOVA and regression. All of these models share the same underlying equation:

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_k X_k + \varepsilon$$

The difference lies in whether the predictors X are categorical (ANOVA),
continuous (regression), or both (ANCOVA / multiple regression).

| Model | Predictors | Typical question |
|---|---|---|
| One-way ANOVA | 1 categorical | Does mean cholesterol differ by race? |
| Two-way ANOVA | 2 categorical | Does it differ by race and gender? |
| Simple linear regression | 1 continuous | Does DBP predict cholesterol? |
| Multiple linear regression | Mixed | What predicts cholesterol after adjustment? |

---

# Setup

```{r packages}
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(GGally))    install.packages("GGally")
if (!require(emmeans))   install.packages("emmeans")
if (!require(car))       install.packages("car")      # VIF, Levene
if (!require(ggpubr))    install.packages("ggpubr")   # Q-Q plots
if (!require(broom))     install.packages("broom")    # tidy model output
if (!require(broom))     install.packages("emmeans")

library(tidyverse)
library(GGally)
library(emmeans)
library(car)
library(ggpubr)
library(broom)
library(emmeans)
```

---

# The dataset

We use the `coronary` dataset from a hypothetical clinical study of 200
participants. After removing missing values, the working dataset contains
64 complete observations across nine variables.

| Variable | Type | Description |
|---|---|---|
| `id` | ID | Subject identifier |
| `cad` | Categorical | Coronary artery disease status: `no cad`, `cad` |
| `sbp` | Continuous | Systolic blood pressure (mmHg) |
| `dbp` | Continuous | Diastolic blood pressure (mmHg) |
| `chol` | Continuous | Total cholesterol (mmol/L) — **outcome** |
| `age` | Continuous | Age (years) |
| `bmi` | Continuous | Body mass index (kg/m²) |
| `race` | Categorical | Ethnicity: `malay`, `chinese`, `indian` |
| `gender` | Categorical | `woman`, `man` |

```{r load-data}
# Load data — adjust path if needed
coronaryNA <- read_csv("coronaryNA.csv")

# Remove missing observations
coronary <- na.omit(coronaryNA)

cat("Rows after removing missing values:", nrow(coronary), "\n")
glimpse(coronary)
```

## Exploratory data analysis

The pairs plot shows histograms on the diagonal, scatter plots with smoothers
below, and Pearson correlations above. Focus on the `chol` row/column to
identify potential predictors.

```{r pairs-plot, fig.height=7, fig.width=8}
GGally::ggpairs(
  coronary %>% select(-id, -cad),
  progress = FALSE,
  upper = list(continuous = wrap("cor", size = 3.5)),
  lower = list(continuous = wrap("smooth", alpha = 0.3, size = 0.5)),
  diag  = list(continuous = wrap("barDiag", bins = 12))
) +
  theme_minimal(base_size = 11) +
  labs(title = "Pairs plot — coronary dataset")
```

```{r descriptives}
# Descriptive statistics by race for the outcome
coronary %>%
  group_by(race) %>%
  summarise(
    n    = n(),
    mean_chol = round(mean(chol), 3),
    sd_chol   = round(sd(chol),   3),
    median    = round(median(chol), 3),
    IQR       = round(IQR(chol), 3)
  )
```

---

# One-way ANOVA

## Research question

Does mean total cholesterol differ across ethnic groups (Malay, Chinese, Indian)?

**Hypotheses:**

- H₀: μ_chinese = μ_indian = μ_malay
- H₁: At least one group mean differs

## Visualise

```{r oneway-plot}
ggplot(coronary, aes(x = race, y = chol, fill = race)) +
  geom_boxplot(alpha = 0.7, width = 0.5, outlier.shape = 21) +
  geom_jitter(width = 0.12, alpha = 0.4, size = 1.5) +
  scale_fill_manual(values = c(
    "chinese" = "#5DCAA5", "indian" = "#F0997B", "malay" = "#FAC775")) +
  labs(title = "Total cholesterol by ethnicity",
       x = NULL, y = "Cholesterol (mmol/L)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## Fit the model

In R, ANOVA is fitted as a linear model using `lm()`. This is intentional — it
reinforces that ANOVA is a special case of linear regression.

```{r oneway-fit}
fit_anova <- lm(chol ~ race, data = coronary)
anova(fit_anova)
```

> **Interpretation:** F(2, 61) = 5.74, p = 0.005. There is a statistically
> significant effect of ethnicity on cholesterol. At least one group differs.
> We proceed to post-hoc pairwise comparisons.

## Post-hoc — Tukey HSD via emmeans

```{r oneway-posthoc}
emm <- emmeans(fit_anova, ~ race)
emm                    # estimated marginal means
pairs(emm)             # Tukey-adjusted pairwise comparisons
```

```{r oneway-emmplot, fig.height=3.5}
plot(emm, comparisons = TRUE) +
  labs(title = "Estimated marginal means with Tukey comparisons",
       x = "Cholesterol (mmol/L)", y = NULL) +
  theme_minimal(base_size = 12)
```

> **Post-hoc results:**
>
> - Chinese vs Indian: estimate = −1.03 mmol/L, p = 0.010 — significant
> - Chinese vs Malay: estimate = −0.21 mmol/L, p = 0.80 — not significant
> - Indian vs Malay: estimate = +0.82 mmol/L, p = 0.022 — significant
>
> Indian participants have significantly higher cholesterol than both Chinese
> and Malay participants.

## Assumption checks

```{r oneway-assumptions, fig.height=4}
# 1. Normality per group (Shapiro-Wilk)
coronary %>%
  group_by(race) %>%
  summarise(W = round(shapiro.test(chol)$statistic, 4),
            p = round(shapiro.test(chol)$p.value,   4))
```

```{r oneway-qq, fig.height=3.5}
# Q-Q plot of residuals
ggqqplot(residuals(fit_anova),
         title = "Q-Q plot — residuals from one-way ANOVA")
```

```{r oneway-levene}
# 2. Homogeneity of variance
leveneTest(chol ~ race, data = coronary)
```

---

# Two-way ANOVA

## Adding gender as a second factor

We extend the model to test both ethnicity and gender simultaneously, first
assuming their effects are **additive** (no interaction).

```{r twoway-fit}
fit_2anova <- lm(chol ~ race + gender, data = coronary)
anova(fit_2anova)
```

> **Results:**
>
> - Race: F(2, 60) = 5.65, p = 0.006 — significant
> - Gender: F(1, 60) = 0.009, p = 0.923 — not significant
>
> Ethnicity remains a significant predictor. After adjusting for ethnicity,
> gender does not explain additional variation in cholesterol.

## Two-way ANOVA with interaction

The additive model assumes the effect of race is identical for men and women.
We test this assumption explicitly by including an interaction term `race * gender`.

```{r interaction-fit}
fit_interaction <- lm(chol ~ race * gender, data = coronary)
anova(fit_interaction)
```

```{r interaction-plot, fig.height=4}
# Visualise potential interaction
coronary %>%
  group_by(race, gender) %>%
  summarise(mean_chol = mean(chol), se = sd(chol)/sqrt(n()), .groups = "drop") %>%
  ggplot(aes(x = race, y = mean_chol,
             colour = gender, group = gender)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_chol - se, ymax = mean_chol + se),
                width = 0.1) +
  scale_colour_manual(values = c("man" = "#378ADD", "woman" = "#D85A30")) +
  labs(title    = "Interaction plot — race × gender on cholesterol",
       subtitle = "Parallel lines suggest no interaction",
       x = NULL, y = "Mean cholesterol (mmol/L)", colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

> **Interaction term:** Race × Gender F(2, 58) = 2.30, p = 0.109 — not
> significant. The effect of ethnicity on cholesterol does not significantly
> differ between men and women. The additive model `chol ~ race + gender` is
> preferred by parsimony.

## Model assumption diagnostics — ggplot2

```{r diag-plots, fig.height=6}
# Extract diagnostics using broom
diag_df <- augment(fit_interaction)

p1 <- ggplot(diag_df, aes(.fitted, .resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "gray50") +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE, colour = "#D85A30", linewidth = 0.8) +
  labs(title = "Residuals vs fitted", x = "Fitted values", y = "Residuals") +
  theme_minimal(base_size = 12)

p2 <- ggqqplot(diag_df$.resid, title = "Normal Q-Q — residuals")

p3 <- ggplot(diag_df, aes(.fitted, sqrt(abs(.std.resid)))) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE, colour = "#D85A30", linewidth = 0.8) +
  labs(title = "Scale-location", x = "Fitted values",
       y = expression(sqrt("|Std. residuals|"))) +
  theme_minimal(base_size = 12)

p4 <- ggplot(diag_df, aes(seq_along(.cooksd), .cooksd)) +
  geom_col(fill = "#378ADD", alpha = 0.7) +
  geom_hline(yintercept = 4 / nrow(coronary), linetype = "dashed",
             colour = "#D85A30") +
  labs(title = "Cook's distance", x = "Observation index",
       y = "Cook's D") +
  theme_minimal(base_size = 12)

gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

> **Diagnostic summary:**
>
> - Residuals vs Fitted: points scatter randomly around zero — linearity OK
> - Q-Q plot: points follow the line closely — normality of residuals satisfied
> - Scale-location: roughly horizontal band — homoscedasticity OK
> - Cook's distance: no observation exceeds the 4/n threshold — no influential outliers

---

# Simple linear regression

Before building a multiple regression model, we examine the relationship
between a single continuous predictor and cholesterol.

## Research question

Does diastolic blood pressure (DBP) predict total cholesterol?

**Hypotheses:**

- H₀: β_dbp = 0 (DBP has no linear association with cholesterol)
- H₁: β_dbp ≠ 0

## Visualise

```{r slr-plot}
ggplot(coronary, aes(x = dbp, y = chol)) +
  geom_point(alpha = 0.5, colour = "#378ADD", size = 2) +
  geom_smooth(method = "lm", colour = "#D85A30", se = TRUE, linewidth = 1) +
  labs(title    = "Simple linear regression — DBP predicting cholesterol",
       subtitle = "Shaded band = 95% confidence interval for the mean",
       x = "Diastolic blood pressure (mmHg)",
       y = "Total cholesterol (mmol/L)") +
  theme_minimal(base_size = 13)
```

## Fit the model

```{r slr-fit}
fit_slr <- lm(chol ~ dbp, data = coronary)
summary(fit_slr)
```

```{r slr-tidy}
tidy(fit_slr, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(., 4)))
```

> **Interpretation:**
>
> - Intercept: predicted cholesterol when DBP = 0 (not clinically meaningful
>   here but retained for model validity)
> - β_dbp: for each 1 mmHg increase in DBP, cholesterol increases by β mmol/L
> - R²: DBP alone explains a modest proportion of cholesterol variability
> - Check p-value for β_dbp to assess statistical significance

## Confidence intervals vs. prediction intervals

```{r slr-intervals}
# New DBP values for prediction
new_dbp <- data.frame(dbp = seq(min(coronary$dbp),
                                 max(coronary$dbp), length.out = 100))

# 95% CI for the mean response (narrower — for the average patient)
ci <- predict(fit_slr, newdata = new_dbp, interval = "confidence")

# 95% PI for an individual new observation (wider — for a single patient)
pi <- predict(fit_slr, newdata = new_dbp, interval = "prediction")

pred_df <- bind_cols(
  new_dbp,
  as.data.frame(ci) %>% rename(fit = fit, ci_lwr = lwr, ci_upr = upr),
  as.data.frame(pi) %>% dplyr::select(pi_lwr = lwr, pi_upr = upr)
)

ggplot(pred_df, aes(x = dbp)) +
  geom_ribbon(aes(ymin = pi_lwr, ymax = pi_upr),
              fill = "#B5D4F4", alpha = 0.4) +
  geom_ribbon(aes(ymin = ci_lwr, ymax = ci_upr),
              fill = "#378ADD", alpha = 0.4) +
  geom_line(aes(y = fit), colour = "#D85A30", linewidth = 1.2) +
  geom_point(data = coronary, aes(x = dbp, y = chol),
             alpha = 0.4, size = 1.5) +
  labs(title    = "Confidence interval vs. prediction interval",
       subtitle = "Dark band = 95% CI for mean response · Light band = 95% PI for individual",
       x = "DBP (mmHg)", y = "Cholesterol (mmol/L)") +
  theme_minimal(base_size = 13)
```

> **Key distinction:**
>
> - The **confidence interval** (dark band) quantifies uncertainty about the
>   *mean* cholesterol at a given DBP — relevant for population estimates
> - The **prediction interval** (light band) quantifies uncertainty for a
>   *single new patient* — always wider because it includes individual variation

## Assumption checks — SLR

```{r slr-assumptions, fig.height=3.5}
# Residuals vs fitted
augment(fit_slr) %>%
  ggplot(aes(.fitted, .resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "gray50") +
  geom_point(alpha = 0.6, colour = "#378ADD") +
  geom_smooth(se = FALSE, colour = "#D85A30", linewidth = 0.8) +
  labs(title = "Residuals vs fitted — SLR",
       x = "Fitted values", y = "Residuals") +
  theme_minimal(base_size = 13)
```

```{r slr-shapiro}
shapiro.test(residuals(fit_slr))
```

---

# Multiple linear regression

## Variable selection — backward stepwise

We begin with a full model including all plausible clinical predictors.
Backward stepwise selection uses the **Akaike Information Criterion (AIC)**
to iteratively remove variables that do not improve model fit.

> **AIC** balances goodness of fit against model complexity.
> Lower AIC = better model. Removing a variable is preferred if it reduces AIC.
> AIC is for variable selection — not hypothesis testing.

```{r mlr-full}
mlr_full <- lm(chol ~ dbp + bmi + race + age + gender, data = coronary)
summary(mlr_full)
```

```{r mlr-backward}
mlr_step <- step(mlr_full, direction = "backward", trace = 1)
```

> After backward selection, the model `chol ~ dbp + race` achieves the lowest
> AIC. Combined with clinical reasoning, this is adopted as the final model.

## Fit the final model

```{r mlr-final}
mlr_final <- lm(chol ~ dbp + race, data = coronary)
summary(mlr_final)
```

```{r mlr-tidy}
tidy(mlr_final, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(., 4)))
```

> **Coefficient interpretation (reference group = Chinese):**
>
> - **Intercept:** predicted cholesterol for a Chinese patient with DBP = 0
>   (not clinically meaningful; needed mathematically)
> - **β_dbp:** for each 1 mmHg increase in DBP, cholesterol increases by
>   0.033 mmol/L, *holding race constant*. p = 0.06 — borderline.
> - **β_raceindian:** Indian patients have 0.88 mmol/L higher cholesterol
>   than Chinese patients, *adjusting for DBP*. p = 0.011 — significant.
> - **β_racemalay:** Malay patients have 0.26 mmol/L higher cholesterol
>   than Chinese patients. p = 0.40 — not significant.
>
> **Model fit:** R² = 0.206 (20.6% of variance explained);
> Adjusted R² = 0.166. The overall model is significant (F(3, 60) = 5.20,
> p = 0.003).

## Multicollinearity — Variance Inflation Factor (VIF)

When predictors are correlated with each other, coefficient estimates become
unstable. VIF quantifies how much variance of each coefficient is inflated
due to collinearity.

- VIF = 1: no collinearity
- VIF 1–5: moderate, acceptable
- VIF > 5–10: high — investigate
- VIF > 10: severe — action required

```{r vif}
vif(mlr_final)
```

> VIF values close to 1 for all predictors indicate no problematic
> multicollinearity in the final model.


## Assumption checks — MLR

```{r mlr-diag, fig.height=6}
diag_mlr <- augment(mlr_final)

p1 <- ggplot(diag_mlr, aes(.fitted, .resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "gray50") +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE, colour = "#D85A30", linewidth = 0.8) +
  labs(title = "Residuals vs fitted", x = "Fitted", y = "Residuals") +
  theme_minimal(base_size = 12)

p2 <- ggqqplot(diag_mlr$.resid, title = "Normal Q-Q — residuals")

p3 <- ggplot(diag_mlr, aes(.fitted, sqrt(abs(.std.resid)))) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE, colour = "#D85A30", linewidth = 0.8) +
  labs(title = "Scale-location", x = "Fitted",
       y = expression(sqrt("|Std. residuals|"))) +
  theme_minimal(base_size = 12)

p4 <- ggplot(diag_mlr, aes(seq_along(.cooksd), .cooksd)) +
  geom_col(fill = "#378ADD", alpha = 0.7) +
  geom_hline(yintercept = 4 / nrow(coronary),
             linetype = "dashed", colour = "#D85A30") +
  annotate("text", x = 5, y = 4 / nrow(coronary) + 0.005,
           label = "4/n threshold", colour = "#D85A30", size = 3, hjust = 0) +
  labs(title = "Cook's distance", x = "Index", y = "Cook's D") +
  theme_minimal(base_size = 12)

gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

```{r mlr-shapiro}
shapiro.test(residuals(mlr_final))
```

> **Diagnostic checklist:**
>
> - Linearity: residuals scatter randomly around zero ✔
> - Normality of residuals: Shapiro-Wilk p > 0.05; Q-Q plot satisfactory ✔
> - Homoscedasticity: scale-location plot shows roughly constant spread ✔
> - Influential observations: no Cook's D exceeds 4/n threshold ✔
> - Multicollinearity: VIF < 5 for all predictors ✔

## Visualise the final model predictions

```{r mlr-pred-plot}
coronary %>%
  mutate(fitted = fitted(mlr_final)) %>%
  ggplot(aes(x = dbp, y = chol, colour = race)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_line(aes(y = fitted), linewidth = 1.1) +
  scale_colour_manual(values = c(
    "chinese" = "#1D9E75", "indian" = "#D85A30", "malay" = "#378ADD")) +
  labs(title    = "Final MLR model — fitted lines by ethnicity",
       subtitle = "Parallel lines reflect the additive (no interaction) model",
       x = "DBP (mmHg)", y = "Cholesterol (mmol/L)", colour = "Ethnicity") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

> Parallel lines confirm the additive assumption: each ethnic group has its
> own intercept (reflecting group differences in baseline cholesterol) but the
> same slope for DBP.

---

# Reporting a linear model

A well-reported linear regression result includes:

1. The regression equation or coefficient table with 95% CIs
2. Overall model significance (F-statistic, df, p-value)
3. R² and adjusted R²
4. Assumption check results
5. Interpretation in clinical terms — not just statistical terms

**Example write-up for the final model:**

> Multiple linear regression was used to examine predictors of total
> cholesterol. After backward stepwise selection, the final model included
> diastolic blood pressure and ethnicity as predictors. The model was
> statistically significant (F(3, 60) = 5.20, p = 0.003) and explained 20.6%
> of the variance in cholesterol (adjusted R² = 0.166). After adjusting for
> DBP, Indian participants had significantly higher cholesterol than Chinese
> participants (β = 0.88, 95% CI [0.22, 1.55], p = 0.011). DBP showed a
> positive but borderline association with cholesterol (β = 0.033,
> 95% CI [−0.001, 0.067], p = 0.06). Regression assumptions were verified:
> residuals were approximately normally distributed (Shapiro-Wilk p > 0.05),
> homoscedasticity was confirmed visually, and VIF values indicated no
> multicollinearity.

---

# Exercises

**Exercise 1 — One-way ANOVA**
Fit a one-way ANOVA testing whether mean SBP differs by ethnicity. Check
assumptions, report the F-statistic and p-value, and run post-hoc comparisons
if warranted.

```{r ex1, eval=FALSE}
# Your code here

```

**Exercise 2 — Two-way ANOVA with interaction**
Fit a two-way ANOVA for SBP with race and gender, including the interaction
term. Produce an interaction plot. Is the interaction significant? What does
this mean clinically?

```{r ex2, eval=FALSE}
# Your code here

```

**Exercise 3 — Simple linear regression**
Fit a simple linear regression of cholesterol on age. Plot the regression
line with confidence and prediction intervals. Interpret the slope and R².

```{r ex3, eval=FALSE}
# Your code here

```

**Exercise 4 — Multiple linear regression**
Fit a full MLR model predicting SBP from age, BMI, gender, and race. Perform
backward stepwise selection. Check VIF on the final model. Report the final
model coefficients with 95% CIs.

```{r ex4, eval=FALSE}
# Your code here

```

---

# Session info

```{r session}
sessionInfo()
```
