---
title: "Confidence Intervals"
author: ''
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
subtitle: "Fundamentals of Biostatistics — Hands-on Session"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,  fig.width  = 5,
  fig.height = 5,
  fig.align  = "center")
```

---

# Setup

```{r packages}
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(dplyr))   install.packages("dplyr")
if (!require(tidyr))   install.packages("tidyr")

library(ggplot2)
library(dplyr)
library(tidyr)
```

---

# What is a confidence interval?

A confidence interval (CI) gives a range of plausible values for a population
parameter, based on sample data.

> **Correct interpretation:** If we repeated this study many times and computed
> a 95% CI each time, 95% of those intervals would contain the true parameter.

> **Common misinterpretation:** "There is a 95% probability that the true
> parameter lies in this specific interval." — This is **wrong**. The parameter
> is fixed; it either is or is not in the interval. Probability applies to the
> *procedure*, not to a single computed interval.

## The general form

$$\text{CI} = \text{point estimate} \pm z^* \times SE$$

Where:

- **Point estimate** — sample statistic (x̄ or p̂)
- **z*** — critical value from the standard normal (1.96 for 95%)
- **SE** — standard error, measuring sampling variability

---

# Simulating the "95% of intervals" interpretation

The best way to understand what 95% confidence means is to simulate it.
We draw 100 samples from a known population and compute a CI from each.

```{r simulation}
set.seed(42)

# True population parameters (known in simulation, unknown in practice)
mu_true <- 120      # true mean SBP (mmHg)
sigma   <- 15       # true SD
n       <- 40       # sample size per study
n_sims  <- 100      # number of simulated studies

# Draw 100 samples and compute 95% CI for each
sims <- data.frame(sim = 1:n_sims) %>%
  rowwise() %>%
  mutate(
    xbar  = mean(rnorm(n, mean = mu_true, sd = sigma)),
    se    = sigma / sqrt(n),
    lower = xbar - 1.96 * se,
    upper = xbar + 1.96 * se,
    captures = (lower <= mu_true & mu_true <= upper)
  ) %>%
  ungroup()

n_captured <- sum(sims$captures)
cat("Number of intervals capturing the true mean:", n_captured, "out of", n_sims, "\n")
cat("Coverage:", round(n_captured / n_sims * 100, 1), "%\n")
```

```{r simulation-plot, fig.height=9}
ggplot(sims, aes(x = sim, y = xbar, colour = captures)) +
  geom_hline(yintercept = mu_true, linetype = "dashed",
             colour = "gray40", linewidth = 0.8) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0, linewidth = 0.6) +
  geom_point(size = 1.5) +
  scale_colour_manual(
    values = c("TRUE" = "#1D9E75", "FALSE" = "#D85A30"),
    labels = c("TRUE" = "Contains true mean", "FALSE" = "Misses true mean")
  ) +
  coord_flip() +
  labs(
    title    = "100 simulated 95% confidence intervals",
    subtitle = paste0("True mean = ", mu_true, " mmHg  ·  n = ", n,
                      " per study  ·  ", n_captured, "/100 intervals capture the true mean"),
    x        = "Simulation",
    y        = "SBP (mmHg)",
    colour   = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

> **Notice:** The true mean (dashed line) is fixed. The intervals move around it.
> Roughly 5 intervals miss — exactly what 95% confidence predicts.

---

# CI for a mean — known σ (z-interval)

Used when the population standard deviation σ is known, or n is large (n > 30).

$$\bar{x} \pm z^* \cdot \frac{\sigma}{\sqrt{n}}$$

## Critical values for common confidence levels

```{r critical-values}
conf_levels <- c(0.90, 0.95, 0.99)
z_stars     <- qnorm(1 - (1 - conf_levels) / 2)

data.frame(
  `Confidence level` = paste0(conf_levels * 100, "%"),
  `α`                = 1 - conf_levels,
  `α/2`              = (1 - conf_levels) / 2,
  `z*`               = round(z_stars, 3)
)
```

## Worked example

```{r z-interval}
# Sample data
x_bar <- 124.3    # sample mean SBP (mmHg)
sigma <- 15       # known population SD
n     <- 40       # sample size

se <- sigma / sqrt(n)
cat("Standard error (SE):", round(se, 3), "mmHg\n\n")

for (cl in c(0.90, 0.95, 0.99)) {
  z_star <- qnorm(1 - (1 - cl) / 2)
  lower  <- x_bar - z_star * se
  upper  <- x_bar + z_star * se
  cat(cl * 100, "% CI: [", round(lower, 1), ",", round(upper, 1), "]\n")
}
```

```{r ci-width-plot}
# Visualise how CI width grows with confidence level
cl_seq <- seq(0.80, 0.99, by = 0.001)
widths <- data.frame(
  conf  = cl_seq,
  width = 2 * qnorm(1 - (1 - cl_seq) / 2) * (sigma / sqrt(n))
)

ggplot(widths, aes(x = conf * 100, y = width)) +
  geom_line(colour = "#378ADD", linewidth = 1.2) +
  geom_vline(xintercept = 95, linetype = "dashed", colour = "#D85A30") +
  annotate("text", x = 95.5, y = min(widths$width) + 1,
           label = "95%", colour = "#D85A30", size = 3.5, hjust = 0) +
  labs(
    title    = "CI width increases with confidence level",
    subtitle = paste0("σ = ", sigma, "  ·  n = ", n),
    x        = "Confidence level (%)",
    y        = "CI width (mmHg)"
  ) +
  theme_minimal(base_size = 13)
```

> **Key insight:** Higher confidence comes at the cost of a wider — less precise — interval.

---

# CI for a mean — unknown σ (t-interval)

In practice, σ is rarely known. We estimate it with the sample SD s and use
the t-distribution, which has heavier tails to account for the added uncertainty.

$$\bar{x} \pm t^*_{n-1} \cdot \frac{s}{\sqrt{n}}$$

The t-distribution converges to the normal as n increases — for n > 30 the
difference is negligible.

## t vs z critical values

```{r t-vs-z}
ns <- c(5, 10, 20, 30, 50, 100, 500)

comparison <- data.frame(
  n      = ns,
  df     = ns - 1,
  t_star = round(qt(0.975, df = ns - 1), 3),
  z_star = 1.960
)

print(comparison)
```

```{r t-vs-z-plot}
n_seq <- 2:100
df_seq <- n_seq - 1

tvals <- data.frame(
  n      = n_seq,
  t_star = qt(0.975, df = df_seq),
  z_star = 1.96
)

ggplot(tvals, aes(x = n)) +
  geom_line(aes(y = t_star, colour = "t*"), linewidth = 1.1) +
  geom_hline(aes(yintercept = 1.96, colour = "z* = 1.96"),
             linetype = "dashed", linewidth = 1) +
  scale_colour_manual(
    values = c("t*" = "#378ADD", "z* = 1.96" = "#D85A30"),
    name   = NULL
  ) +
  labs(
    title    = "t* converges to z* = 1.96 as n increases",
    subtitle = "95% confidence level",
    x        = "Sample size (n)",
    y        = "Critical value"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## Worked example using t.test()

```{r t-interval}
set.seed(7)
sbp_sample <- rnorm(25, mean = 124, sd = 16)   # simulated sample of 25 patients

# Summary statistics
cat("n   =", length(sbp_sample), "\n")
cat("x̄   =", round(mean(sbp_sample), 2), "mmHg\n")
cat("s   =", round(sd(sbp_sample), 2),   "mmHg\n")
cat("SE  =", round(sd(sbp_sample) / sqrt(length(sbp_sample)), 2), "mmHg\n\n")

# 95% CI using t.test
result <- t.test(sbp_sample, conf.level = 0.95)
cat("95% CI: [",
    round(result$conf.int[1], 1), ",",
    round(result$conf.int[2], 1), "] mmHg\n")
```

---

# CI for a proportion

When the outcome is binary (yes/no, event/no event), we estimate a proportion p̂
and construct a CI around it.

$$\hat{p} \pm z^* \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

Valid when n·p̂ ≥ 5 and n·(1−p̂) ≥ 5 (normal approximation holds).

## Worked example

```{r prop-ci}
x <- 87     # number of events (e.g., patients who responded)
n <- 250    # total sample size

p_hat <- x / n
se_p  <- sqrt(p_hat * (1 - p_hat) / n)
z_star <- qnorm(0.975)

lower <- p_hat - z_star * se_p
upper <- p_hat + z_star * se_p

cat("p̂  =", round(p_hat, 3), "\n")
cat("SE =", round(se_p, 4),  "\n")
cat("95% CI: [", round(lower, 3), ",", round(upper, 3), "]\n")
cat("95% CI: [",
    paste0(round(lower * 100, 1), "%"), ",",
    paste0(round(upper * 100, 1), "%"), "]\n")

# Using prop.test for comparison (continuity-corrected)
prop.test(x, n, conf.level = 0.95)
```

## How CI width varies with p̂

```{r prop-ci-width}
p_seq <- seq(0.01, 0.99, by = 0.01)
n_vals <- c(50, 100, 250, 500)

prop_widths <- expand.grid(p = p_seq, n = n_vals) %>%
  mutate(
    se    = sqrt(p * (1 - p) / n),
    width = 2 * 1.96 * se
  )

ggplot(prop_widths, aes(x = p, y = width, colour = factor(n), group = factor(n))) +
  geom_line(linewidth = 1) +
  scale_colour_manual(
    values = c("50"  = "#D85A30", "100" = "#BA7517",
               "250" = "#378ADD", "500" = "#1D9E75"),
    name   = "n"
  ) +
  labs(
    title    = "CI width for a proportion",
    subtitle = "95% confidence level — width maximised at p̂ = 0.5",
    x        = "Estimated proportion (p̂)",
    y        = "CI width"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")
```

> **Key insight:** CI width is maximised when p̂ = 0.5. This is why sample size
> formulas for proportions use p = 0.5 as the conservative default.

---

# What affects CI width?

Three factors control how wide your interval is:

```{r width-factors}
# Base scenario
n_base     <- 100
sigma_base <- 15
z_base     <- qnorm(0.975)   # 95% CI

base_width <- 2 * z_base * sigma_base / sqrt(n_base)
cat("Base CI width (n=100, σ=15, 95%):", round(base_width, 2), "mmHg\n\n")

# Factor 1: double n
w_2n <- 2 * z_base * sigma_base / sqrt(200)
cat("Double n (n=200):          width =", round(w_2n, 2),
    " | reduction:", round((1 - w_2n / base_width) * 100, 0), "%\n")

# Factor 2: double σ
w_2s <- 2 * z_base * (sigma_base * 2) / sqrt(n_base)
cat("Double σ (σ=30):           width =", round(w_2s, 2),
    " | increase: ", round((w_2s / base_width - 1) * 100, 0), "%\n")

# Factor 3: increase to 99% CI
w_99 <- 2 * qnorm(0.995) * sigma_base / sqrt(n_base)
cat("99% CI (z*=2.576):         width =", round(w_99, 2),
    " | increase: ", round((w_99 / base_width - 1) * 100, 0), "%\n")
```

```{r width-summary-plot}
scenarios <- data.frame(
  scenario = c("Base\n(n=100, σ=15, 95%)",
               "Double n\n(n=200)",
               "Double σ\n(σ=30)",
               "99% CI\n(z*=2.576)"),
  width = c(base_width, w_2n, w_2s, w_99),
  group = c("base", "n", "sigma", "conf")
)

ggplot(scenarios, aes(x = reorder(scenario, width), y = width,
                      fill = group)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_text(aes(label = round(width, 1)), hjust = -0.2, size = 4) +
  scale_fill_manual(
    values = c("base"  = "#888780", "n"     = "#1D9E75",
               "sigma" = "#D85A30", "conf"  = "#378ADD")
  ) +
  coord_flip() +
  ylim(0, max(scenarios$width) * 1.15) +
  labs(
    title = "Effect of each factor on CI width",
    x     = NULL,
    y     = "CI width (mmHg)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```


---

# Exercises

**Exercise 1.**
A sample of 36 patients has a mean resting heart rate of 74 bpm with a known
population SD of 10 bpm. Construct 90%, 95%, and 99% confidence intervals.
Interpret the 95% interval in one sentence.

```{r ex1, eval=FALSE}
# Your code here

```

**Exercise 2.**
In a clinical trial of 180 patients, 54 experienced a side effect. Compute
the 95% CI for the true proportion experiencing the side effect. Check that
the normal approximation conditions are met.

```{r ex2, eval=FALSE}
# Your code here

```

**Exercise 3.**
Reproduce the simulation from Section 2 but change n to 15. How does the
coverage change, and why? What happens if you use z* = 1.96 instead of t*?

```{r ex3, eval=FALSE}
# Your code here

```


---

# Session info

```{r session}
sessionInfo()
```
