---
title: "Categorical Data Analysis"
author: "Statistical Methods in R"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
subtitle: "Relative Risk, Odds Ratio, Chi-Square, Fisher's Exact Test & Goodness-of-Fit"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width  = 7,
  fig.height = 5
)
```

---

# Overview

This notebook covers the core methods for analyzing **categorical data** — data that takes a limited set of discrete values (categories) rather than continuous measurements. We focus on:

| Method | Purpose |
|---|---|
| Relative Risk (RR) | Compare risk of an outcome between two groups |
| Odds Ratio (OR) | Compare odds of an outcome between two groups |
| Chi-Square Test for Independence | Test association between two categorical variables |
| Fisher's Exact Test | Exact test for small samples / sparse tables |
| Chi-Square Goodness-of-Fit | Test whether data follows a hypothesized distribution |

**No regression models are included** — these topics are covered in a separate notebook.

---

# Setup

```{r packages}
# Install packages if not already available
if (!requireNamespace("epitools",  quietly = TRUE)) install.packages("epitools")
if (!requireNamespace("vcd",       quietly = TRUE)) install.packages("vcd")
if (!requireNamespace("ggplot2",   quietly = TRUE)) install.packages("ggplot2")
if (!requireNamespace("dplyr",     quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("knitr",     quietly = TRUE)) install.packages("knitr")

library(epitools)   # riskratio(), oddsratio()
library(vcd)        # mosaic plots, assocstats()
library(ggplot2)    # visualizations
library(dplyr)      # data manipulation
library(knitr)      # kable() for nice tables
```

---

# The Study Scenario

Throughout this notebook we use a simulated **cohort study** examining the association between **smoking status** and **lung disease diagnosis** among 300 participants.

```{r dataset}
# ── Simulated data ────────────────────────────────────────────────────────────
set.seed(42)
n <- 300

smoking <- sample(c("Smoker", "Non-smoker"), n,
                  replace = TRUE, prob = c(0.45, 0.55))

# Disease probability: higher in smokers
p_disease <- ifelse(smoking == "Smoker", 0.40, 0.10)
disease    <- rbinom(n, 1, p_disease)
disease    <- factor(disease, levels = c(1, 0),
                     labels = c("Disease", "No Disease"))
smoking    <- factor(smoking, levels = c("Smoker", "Non-smoker"))

study_data <- data.frame(smoking, disease)

head(study_data, 10)
```

```{r summary-table}
# Frequency table
tab <- table(Smoking = study_data$smoking,
             Disease = study_data$disease)

tab
```

```{r proportions}
# Row proportions (risk of disease within each group)
kable(round(prop.table(tab, margin = 1), 3),
      caption = "Row Proportions (risk within each smoking group)")
```

---

# Chi-Square Test for Independence

## Concept

Tests whether two categorical variables are **statistically independent** — i.e., whether the distribution of one variable differs across levels of the other.

$$H_0: \text{Smoking and disease are independent}$$
$$H_1: \text{Smoking and disease are associated}$$

**Test statistic:**

$$\chi^2 = \sum_{i,j} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}$$

where $E_{ij} = \frac{(\text{Row total}_i) \times (\text{Col total}_j)}{N}$

**Degrees of freedom:** $(r - 1)(c - 1)$

> **Assumptions:**
> - Observations are independent
> - Expected cell counts ≥ 5 in more than 75% (otherwise use Fisher's Exact Test)


## Using `chisq.test()`

```{r chi-sq-builtin}
chi_test <- chisq.test(tab)
print(chi_test)

chi_test$statistic

# Inspect expected counts
cat("\nExpected cell counts:\n")
print(round(chi_test$expected, 2))

# Standardized residuals
cat("\nStandardized (Pearson) residuals:\n")
print(round(chi_test$residuals, 3))
```
X-squared = 18.245, df = 1, p-value = 1.943e-05, there is a relationship between smoking and disease. 


## Visualizing the Association

```{r chi-sq-barplot}
# Grouped bar chart
ggplot(study_data, aes(x = smoking, fill = disease)) +
  geom_bar(position = "dodge")
```

## Effect Size: Cramér's V

The chi-square statistic depends on sample size. **Cramér's V** standardizes it into a measure of association strength (0 = none, 1 = perfect):

$$V = \sqrt{\frac{\chi^2}{N \cdot \min(r-1,\, c-1)}}$$

```{r cramers-v}
# Effect size V

assocstats(tab)
```

---

# Fisher's Exact Test

## Concept

Fisher's Exact Test computes the **exact** probability of observing data as extreme as (or more extreme than) what was observed, under H₀ of independence — without relying on large-sample approximations.

**When to use:**
- Small samples
- Any expected cell count < 5
- As a general-purpose alternative when you want exact p-values

The test is based on the **hypergeometric distribution**:

$$P(\text{table} \mid \text{marginals fixed}) = \frac{\binom{n_1}{a}\binom{n_0}{c}}{\binom{N}{a+c}}$$

## Demonstration with a Sparse Table

```{r fisher-sparse}
# Small study: new drug vs. placebo, rare side effect
sparse_tab <- matrix(
  c(3, 1,   # Drug:    3 with side effect, 1 without
    0, 9),  # Placebo: 0 with side effect, 9 without
  nrow = 2, byrow = TRUE,
  dimnames = list(
    Treatment = c("Drug", "Placebo"),
    SideEffect = c("Yes", "No")
  )
)

kable(addmargins(sparse_tab),
      caption = "Sparse 2×2 Table: Drug Trial (n = 13)")
```

```{r chi-sq-vs-fisher}
# Check expected counts — chi-square unreliable here
cat("Expected counts under chi-square:\n")
print(round(chisq.test(sparse_tab)$expected, 2))
cat("\nWarning: several expected counts < 5 → use Fisher's Exact Test\n")
```

```{r fisher-exact}
fisher_result <- fisher.test(sparse_tab)
print(fisher_result)

cat(sprintf("\nExact p-value: %.4f\n",       fisher_result$p.value))
cat(sprintf("Odds Ratio:    %.3f\n",         fisher_result$estimate))
cat(sprintf("95%% CI:        %.3f – %.3f\n",
            fisher_result$conf.int[1], fisher_result$conf.int[2]))
```
---

# Session Information

```{r session-info}
sessionInfo()
```
