---
title: 'Multiple Logistic Regression: Analyzing Cancer Risk'
author: "BioData Analysis Tutorial"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 5
)
```

---

# Introduction

**Logistic regression** is a statistical method used to model the relationship between one or more independent variables (predictors) and a **binary outcome** — an outcome with only two possible categories, usually coded as **0** and **1**.

Unlike linear regression (where the outcome is continuous), logistic regression predicts the **probability** that an event occurs — for example, whether a patient **does** or **does not** have cancer.

Because probabilities must lie between 0 and 1, logistic regression transforms the relationship using the **logit function**, which maps probabilities onto the entire real line:

$$\text{logit}(p) = \log\!\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k$$

The model **does not predict probability directly**, but predicts the **log odds**, which can then be transformed back:

$$\hat{p} = \frac{e^{\beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k}}{1 + e^{\beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k}}$$

---

## Study Design

Suppose you're a researcher studying factors associated with a positive cancer diagnosis. You collect data on **200 patients**. For each patient, you know:

| Variable | Type | Description |
|---|---|---|
| `Age` | Continuous | Patient age in years |
| `BMI` | Continuous | Body Mass Index |
| `SmokingStatus` | Categorical | Smoker / Non-Smoker |
| `FamilyHistory` | Categorical | Family history of cancer: Yes / No |
| `CancerStatus` | Binary outcome | 1 = diagnosed, 0 = not diagnosed |

---

# Step 1: Load Packages

```{r load-packages}

# Install missing packages and load libraries

packages <- c("tidyverse", "car", "pscl", "ResourceSelection", "pROC")

installed <- installed.packages()[, "Package"]
missing   <- packages[!packages %in% installed]

if (length(missing) > 0) {
  message("Installing missing packages: ", paste(missing, collapse = ", "))
  install.packages(missing)
} else {
  message("All packages already installed.")
}

library(tidyverse)
library(car)
library(pscl)
library(ResourceSelection)
library(pROC)

```


# Load the data
```{r data-summary}

data <- read_csv("data_cancer.csv")
# Quick overview
glimpse(data)

# Outcome distribution
data %>%
  count(CancerStatus) %>%
  mutate(Proportion = round(n / sum(n), 2))

```

---

# Step 2: Fit the Multiple Logistic Regression Model

```{r fit-model}
# Fit the logistic regression model
model <- glm(CancerStatus ~ Age + BMI + SmokingStatus + FamilyHistory,
             data   = data,
             family = binomial)

# View the model summary
summary(model)
```

---

# Step 3: Interpret the Results

## Odds Ratios

Coefficients are on the **log-odds scale**. Exponentiating converts them to **odds ratios (OR)** — the most intuitive unit of interpretation.

```{r odds-ratios}
# Compute odds ratios with 95% confidence intervals
OR_table <- exp(cbind(OR = coef(model), confint(model)))
round(OR_table, 3)
```

**Interpretation of each predictor:**

- **Age** (OR ≈ 1.059): Each additional year of age is associated with ~**5.9% higher odds** of cancer, holding other factors constant.
- **BMI** (OR ≈ 1.157): Each unit increase in BMI raises the odds by ~**15.7%**.
- **SmokingStatus — Smoker** (OR ≈ 4.25): Smokers have about **4.25 × the odds** of cancer compared to non-smokers.
- **FamilyHistory — Yes** (OR ≈ 6.58): Patients with a family history have over **6.58 × the odds** compared to those without — the strongest predictor in the model.

All predictors are statistically significant at the 0.05 level.

## Coefficient Plot

```{r coef-plot}
# Build a tidy OR data frame for plotting
or_df <- data.frame(
  Term     = rownames(OR_table),
  OR       = OR_table[, "OR"],
  Lower    = OR_table[, "2.5 %"],
  Upper    = OR_table[, "97.5 %"]
) %>%
  filter(Term != "(Intercept)")

ggplot(or_df, aes(x = OR, y = reorder(Term, OR))) +
  geom_point(size = 4, color = "tomato") +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper), height = 0.25, color = "tomato") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  labs(title = "Odds Ratios with 95% Confidence Intervals",
       x = "Odds Ratio (OR)", y = NULL) +
  theme_minimal(base_size = 13)
```

---

# Step 4: Visualize Model Performance (ROC & AUC)

The **Receiver Operating Characteristic (ROC) curve** shows how well the model discriminates between patients with and without cancer across all possible thresholds.

```{r roc-curve}
# Get predicted probabilities from the model
probabilities <- predict(model, type = "response")

# Actual outcomes
actual <- data$CancerStatus

# Create ROC object
roc_obj <- roc(actual, probabilities, quiet = TRUE)

# Plot ROC curve
plot(roc_obj,
     main = "ROC Curve for Cancer Prediction",
     col  = "#2C7BB6",
     lwd  = 3)
legend("bottomright",
       legend = paste0("AUC = ", round(auc(roc_obj), 3)),
       col = "#2C7BB6", lwd = 3, bty = "n")

# Print AUC
cat("Area under the curve (AUC):", round(auc(roc_obj), 3), "\n")
```

**AUC Interpretation guide:**

| AUC | Discrimination |
|---|---|
| 0.5 | No discrimination (random) |
| 0.7 – 0.8 | Acceptable |
| 0.8 – 0.9 | Good |
| > 0.9 | Excellent |

An AUC of **~0.799** indicates the model can reliably rank patients by cancer risk **79.9% of the time**.

---

# Step 6: Confusion Matrix & Classification Metrics

Choose a **threshold** (default = 0.5) to convert predicted probabilities into class labels.

```{r confusion-matrix}
# Classify patients using 0.5 threshold
predicted_class <- ifelse(probabilities > 0.5, 1, 0)

# Confusion matrix
cm <- table(Predicted = predicted_class, Actual = actual)
print(cm)
```

```{r classification-metrics}
# Accuracy
accuracy    <- mean(predicted_class == actual)

# Sensitivity (True Positive Rate — how well we catch cancer cases)
sensitivity <- sum(predicted_class == 1 & actual == 1) / sum(actual == 1)

# Specificity (True Negative Rate — how well we rule out cancer)
specificity <- sum(predicted_class == 0 & actual == 0) / sum(actual == 0)

cat(sprintf("Accuracy    : %.2f\n", accuracy))
cat(sprintf("Sensitivity : %.2f\n", sensitivity))
cat(sprintf("Specificity : %.2f\n", specificity))
```

**Interpretation:**

- **True Negatives**: Patients correctly classified as no cancer.
- **True Positives**: Patients correctly classified as having cancer.
- **False Negatives**: Cancer cases the model missed *(clinically costly)*.
- **False Positives**: Non-cancer patients flagged as at risk.


---

# Step 6: Check Assumptions (Diagnostics)

## 7.1 Linearity in the Logit (Box-Tidwell Test)

Check whether continuous predictors have a linear relationship with the log-odds by adding interaction terms $x \cdot \log(x)$.

```{r box-tidwell}
# Add log-transformed terms
data$logAge <- log(data$Age)
data$logBMI <- log(data$BMI)

# Fit model with Box-Tidwell interaction terms
model_bt <- glm(CancerStatus ~ Age + BMI + SmokingStatus + FamilyHistory +
                  I(Age * logAge) + I(BMI * logBMI),
                data   = data,
                family = binomial)

summary(model_bt)
```

> If `I(Age * logAge)` and `I(BMI * logBMI)` are **not significant**, the linearity assumption holds. ✔

## 6.2 Multicollinearity (VIF)

```{r vif}
# Variance Inflation Factors
vif(model)
```

> VIF values **< 5** indicate acceptable multicollinearity. Values near 1–2 are ideal.

## 6.3 Influential Points (Cook's Distance)

```{r cooks-distance}
cooksd <- cooks.distance(model)

plot(cooksd,
     pch  = 19,
     cex  = 1.2,
     col  = ifelse(cooksd > 4 / nrow(data), "tomato", "steelblue"),
     main = "Cook's Distance — Influential Observations",
     ylab = "Cook's Distance")
abline(h = 4 / nrow(data), col = "red", lty = 2)

# Identify influential points
influential <- which(cooksd > 4 / nrow(data))
cat("Number of influential points:", length(influential), "\n")
if (length(influential) > 0) cat("Row indices:", influential, "\n")
```

---

# Step 7: Goodness-of-Fit

## 7.1 McFadden's Pseudo R²

```{r pseudo-r2}
pR2(model)
```

> McFadden's R² values of **0.2–0.4** suggest a reasonably well-fitting model.

## 7.2 Likelihood Ratio Test

Compare the fitted model to a **null model** (intercept only) to confirm that the predictors improve fit.

```{r lrt}
# Null (intercept-only) model
model_null <- glm(CancerStatus ~ 1, data = data, family = binomial)

# Likelihood Ratio Test
anova(model_null, model, test = "Chisq")
```

> A **significant p-value** means the predictors collectively improve the model beyond the null. ✔

## 7.3 Hosmer-Lemeshow Test

```{r hosmer-lemeshow}
# Hosmer-Lemeshow goodness-of-fit test (g = 10 groups)
hoslem.test(data$CancerStatus, fitted(model), g = 10)
```

> A **non-significant p-value (> 0.05)** indicates good calibration — predicted probabilities match observed event rates.



---

# Part II — Poisson Regression

## The `epil` Dataset

The `epil` dataset (Thall & Vail, 1990) records **epileptic seizure counts** from a randomized placebo-controlled trial of the anti-epileptic drug **progabide**. Fifty-nine patients were followed over **four consecutive two-week periods**.

```{r epil-load}
data(epil, package = "MASS")
str(epil)
```

```{r epil-dict}
epi_dict <- data.frame(
  Variable = c("y","trt","base","age","V4","subject","period","lbase","lage"),
  Description = c(
    "Seizure count in a 2-week period — the count outcome",
    "Treatment group: 'placebo' or 'progabide'",
    "Baseline seizure count (8 weeks before trial)",
    "Patient age (years)",
    "Indicator: 4th observation period (captures period-specific shift)",
    "Patient ID (59 patients × 4 periods = 236 observations)",
    "Observation period: 1 to 4",
    "log(base / 4) — log-scaled baseline rate per 2-week period",
    "log(age) — log-transformed patient age"
  )
)

```

## Exploratory Data Analysis

```{r epil-summary}
cat("Seizure count summary:\n")
summary(epil$y)
cat(sprintf("\n  Mean:     %.2f\n  Variance: %.2f\n  Ratio (Var/Mean): %.2f\n",
            mean(epil$y), var(epil$y), var(epil$y) / mean(epil$y)))
cat("\nNote: Variance >> Mean suggests overdispersion — a key diagnostic concern.\n")
```

```{r epil-dist-plot}
ggplot(epil, aes(x = y, fill = trt)) +
  geom_histogram(binwidth = 2, color = "white", alpha = 0.75) +
  scale_fill_manual(values = c("placebo" = "#2c7fb8", "progabide" = "#d7301f")) +
  facet_wrap(~ trt, ncol = 2,
             labeller = labeller(trt = c(placebo = "Placebo", progabide = "Progabide"))) +
  labs(
    title    = "Distribution of Seizure Counts by Treatment Group",
    subtitle = "Right-skewed counts with heavy tails — characteristic of overdispersion",
    x = "Seizure count (per 2-week period)", y = "Frequency"
  ) +
  theme(legend.position = "none")
```

```{r epil-baseline}
ggplot(epil, aes(x = base, y = y, color = trt)) +
  geom_point(alpha = 0.35) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("placebo" = "#2c7fb8", "progabide" = "#d7301f"),
                     labels = c("Placebo","Progabide")) +
  scale_x_log10() + scale_y_log10() +
  labs(
    title    = "Baseline Seizures vs. Observed Count (log-log scale)",
    subtitle = "Strong positive relationship — baseline is a critical confounder",
    x = "Baseline seizure count (log scale)",
    y = "Period seizure count (log scale)", color = "Treatment"
  ) +
  theme(legend.position = "top")
```

## Model Building

### Model 1 — Null

```{r epil-null}
ep0 <- glm(y ~ 1, data = epil, family = poisson)
cat("Null deviance:", round(deviance(ep0), 2), "on", df.residual(ep0), "df\n")
cat("Global mean count:", round(mean(epil$y), 3), "\n")
```

### Model 2 — Treatment Only (Unadjusted)

```{r epil-trt}
ep1 <- glm(y ~ trt, data = epil, family = poisson)
summary(ep1)

cat(sprintf("\nUnadjusted IRR for progabide: %.3f  (95%% CI: %.3f – %.3f)\n",
            exp(coef(ep1)["trtprogabide"]),
            exp(confint.default(ep1)["trtprogabide", 1]),
            exp(confint.default(ep1)["trtprogabide", 2])))
```

### Model 3 — Adjusted (Treatment + Baseline + Age + Period)

```{r epil-adjusted}
ep2 <- glm(y ~ trt + lbase + lage + V4,
           data   = epil,
           family = poisson)
summary(ep2)
```

```{r}
IRR_table <- exp(cbind(
  IRR      = coef(ep2),
  Lower95  = confint(ep2)[, 1],
  Upper95  = confint(ep2)[, 2]
))

cat("\nIRR with 95% CI:\n")
print(round(IRR_table, 3))
```

Interpretation: 

Reporting Incidence Rate Ratios (IRR). IRR = 1 means no effect; IRR > 1 means higher rate; IRR < 1 means lower rate.

Term-by-Term Interpretation
Intercept (IRR = 5.740) - The baseline expected seizure count (when all predictors = 0) is 5.74 per period. Highly significant (p < 0.001), but not meaningful on its own.


trtprogabide (IRR = 0.983, p = 0.715) - Patients on Progabide have a seizure rate 1.7% lower than the control group. The 95% CI (0.894 – 1.080) crosses 1, meaning the effect could go either direction. Not statistically significant (p = 0.715) — no strong evidence that Progabide reduces seizures compared to placebo.


lbase — log(baseline seizure count) (IRR = 3.405, p < 0.001) - The strongest predictor in the model. A one-unit increase in log-baseline seizures is associated with 3.4× higher seizure rate during the trial. Narrow CI (3.195 – 3.629) and p < 0.001 → highly significant.
Makes clinical sense: patients with more seizures at baseline tend to have more during treatment.


lage — log(age) (IRR = 1.800, p < 0.001) - A one-unit increase in log-age is associated with 1.8× higher seizure rate. CI (1.451 – 2.233) does not cross 1 → statistically significant.
Older patients tend to have higher seizure rates, after adjusting for other factors.


V4 — Visit 4 indicator (IRR = 0.851, p = 0.003) - At Visit 4, the seizure rate is 14.9% lower compared to the reference visit. CI (0.765 – 0.947) does not cross 1 → statistically significant.
Suggests a time effect — seizure rates decline by the 4th visit, possibly due to treatment adaptation or natural variability. 




#Plot the results 

```{r epil-irr-forest}
cf_ep2 <- data.frame(
  Term     = names(coef(ep2))[-1],
  IRR      = exp(coef(ep2))[-1],
  CI_lower = exp(confint.default(ep2))[-1, 1],
  CI_upper = exp(confint.default(ep2))[-1, 2],
  p_value  = coef(summary(ep2))[-1, 4]
)

ggplot(cf_ep2, aes(x = IRR, y = reorder(Term, IRR))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper),
                 height = 0.25, color = "#2c7fb8", linewidth = 0.9) +
  geom_point(aes(color = p_value < 0.05), size = 3.5) +
  scale_color_manual(
    values = c("TRUE" = "#d7301f", "FALSE" = "#2c7fb8"),
    labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05"), name = NULL
  ) +
  scale_x_log10() +
  labs(
    title    = "Incidence Rate Ratios: Epileptic Seizures (Adjusted Poisson Model)",
    subtitle = "IRR < 1 = fewer seizures; IRR > 1 = more seizures",
    x = "Incidence Rate Ratio (log scale)", y = NULL
  ) +
  theme(legend.position = "top")
```


## Model Selection

### Likelihood Ratio Tests

```{r epil-lrt}
anova(ep0, ep1, ep2, test = "LRT") 
```

### AIC / BIC Comparison

```{r epil-aic}
poisson_comp <- data.frame(
  Model    = c("Null", "Treatment only",
               "Adjusted (trt + lbase + lage + V4)"),
  k        = c(length(coef(ep0)), length(coef(ep1)),
               length(coef(ep2))),
  Deviance = round(c(deviance(ep0), deviance(ep1), deviance(ep2)), 2),
  AIC      = round(c(AIC(ep0), AIC(ep1), AIC(ep2)), 2),
  BIC      = round(c(BIC(ep0), BIC(ep1), BIC(ep2)), 2)
)

knitr::kable(poisson_comp, caption = "Poisson Model Comparison")

```

**Selected model:** `ep2` (adjusted). 

## Diagnostics

### 1. Overdispersion Check

```{r epil-overdispersion}
phi <- deviance(ep2) / df.residual(ep2)
p_formal <- pchisq(deviance(ep2), df.residual(ep2), lower.tail = FALSE)

cat("── Overdispersion Diagnostic ──────────────────────────────────\n")
cat(sprintf("Residual deviance:  %.2f\n", deviance(ep2)))
cat(sprintf("Residual df:        %d\n",   df.residual(ep2)))
cat(sprintf("Dispersion φ = Dev/df: %.3f\n\n", phi))
cat(sprintf("Formal chi-square test p-value: %.8f\n\n", p_formal))
cat(sprintf("Interpretation: φ = %.2f >> 1 → strong overdispersion detected.\n", phi))
cat("Standard Poisson SEs are underestimated → inference is anti-conservative.\n")
cat("Remedies: Quasi-Poisson (corrects SE) or Negative Binomial (re-specifies variance).\n")
```

```{r epil-meanvar}
# Mean-variance plot by subject
mean_var <- epil |>
  group_by(subject) |>
  summarise(subj_mean = mean(y), subj_var = var(y), .groups = "drop")

ggplot(mean_var, aes(x = subj_mean, y = subj_var)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "#2c7fb8", linewidth = 0.9) +
  geom_point(color = "#2c7fb8", alpha = 0.65, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#d7301f", linewidth = 1) +
  annotate("text", x = 8,  y = 2,
           label = "Poisson: Var = Mean", color = "#2c7fb8", size = 3.5) +
  annotate("text", x = 23, y = 100,
           label = "Observed trend", color = "#d7301f", size = 3.5) +
  labs(
    title    = "Mean-Variance Relationship by Subject",
    subtitle = "Observed variance greatly exceeds the mean — overdispersion confirmed",
    x = "Subject mean seizure count", y = "Subject variance"
  )
```

### 2. QQ Plot of Deviance Residuals

```{r epil-qq}
qqnorm(residuals(ep2, type = "deviance"),
       main  = "QQ Plot of Deviance Residuals (Poisson Model)",
       col   = "#2c7fb8", pch = 16, cex = 0.7)
qqline(residuals(ep2, type = "deviance"),
       col = "#d7301f", lwd = 1.5)
mtext("Heavy tails confirm overdispersion — Poisson assumption is violated",
      side = 3, cex = 0.8, col = "gray40")
```

## Remedying Overdispersion

### Quasi-Poisson: Corrected Standard Errors

```{r epil-quasi}
ep2_quasi <- glm(y ~ trt + lbase + lage + V4,
                 data   = epil,
                 family = quasipoisson)
summary(ep2_quasi)
```



# Session Info

```{r session-info}
sessionInfo()
```
