---
title: "Parametric & Non-Parametric Tests"
author: ''
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
subtitle: "Fundamentals of Biostatistics — Hands-on Session"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4)
```

---

# Setup

```{r packages}
if (!require(ggplot2))  install.packages("ggplot2")
if (!require(dplyr))    install.packages("dplyr")
if (!require(tidyr))    install.packages("tidyr")
if (!require(car))      install.packages("car")       # Levene's test
if (!require(rstatix))  install.packages("rstatix")   # effect sizes, pipe-friendly tests
if (!require(ggpubr))   install.packages("ggpubr")    # Q-Q plots, publication figures
if (!require(coin))     install.packages("coin")       # exact non-parametric tests
if (!require(FSA))      install.packages("FSA")        # Dunn post-hoc test

library(ggplot2)
library(dplyr)
library(tidyr)
library(car)
library(rstatix)
library(ggpubr)
library(coin)
library(FSA)
```

---

# One-sample t-test

## Clinical scenario

Resting systolic blood pressure (SBP) in healthy adults is established at
**120 mmHg**. A clinician suspects that patients with early-stage chronic kidney
disease (CKD) have elevated SBP even before diagnosis. A sample of 30 CKD
patients is measured at their first clinic visit.

**Hypotheses:**
- H₀: μ = 120 mmHg (CKD patients have normal SBP)
- H₁: μ ≠ 120 mmHg (two-tailed)

```{r one-sample-data}
#simulate the data

set.seed(42)
sbp_ckd <- round(rnorm(30, mean = 128, sd = 14), 1)

cat("n    =", length(sbp_ckd), "\n")
cat("Mean =", round(mean(sbp_ckd), 2), "mmHg\n")
cat("SD   =", round(sd(sbp_ckd),   2), "mmHg\n")
cat("SE   =", round(sd(sbp_ckd) / sqrt(length(sbp_ckd)), 2), "mmHg\n")
```

## Assumption check — normality

For a one-sample t-test with n = 30, normality of the outcome is required.

```{r one-sample-normality, fig.height=3.5}
# Shapiro-Wilk test (sensitive to departures from normality; best for n < 50)
shapiro.test(sbp_ckd)

# Q-Q plot
ggqqplot(sbp_ckd, title = "Q-Q plot — SBP in CKD patients") +
  labs(x = "Theoretical quantiles", y = "Sample quantiles (mmHg)")
```

> **Interpretation:** p > 0.05 on Shapiro-Wilk and points close to the line on
> the Q-Q plot → normality assumption is satisfied → proceed with t-test.

## Run the test

```{r one-sample-test}
result_1s <- t.test(sbp_ckd, mu = 120, alternative = "two.sided")
print(result_1s)
```

## Effect size — Cohen's d

```{r one-sample-d}
cohens_d <- (mean(sbp_ckd) - 120) / sd(sbp_ckd)
cat("Cohen's d =", round(cohens_d, 3),
    " | Interpretation:",
    ifelse(abs(cohens_d) < 0.2, "negligible",
    ifelse(abs(cohens_d) < 0.5, "small",
    ifelse(abs(cohens_d) < 0.8, "medium", "large"))), "\n")
```

## Interpret and report

```{r one-sample-report}
cat("Mean SBP in CKD patients:", round(mean(sbp_ckd), 1), "mmHg\n")
cat("95% CI: [",
    round(result_1s$conf.int[1], 1), ",",
    round(result_1s$conf.int[2], 1), "] mmHg\n")
cat("t(", result_1s$parameter, ") =",
    round(result_1s$statistic, 3),
    ", p =", round(result_1s$p.value, 4), "\n")
```

> **Conclusion:** Mean SBP in early CKD patients was significantly higher than
> the reference value of 120 mmHg (t(29) = 2.793, p < 0.05, Cohen's d = 0.51, medium).
> This suggests subclinical hypertension may be present at CKD diagnosis.

## Non-parametric analogue — Wilcoxon signed-rank test (one-sample)

Use when the normality assumption is violated. Instead of testing whether
the **mean** equals a reference value, this test asks whether the **median**
of the population equals the reference value. It works by ranking the absolute
deviations from the reference and checking whether positive and negative ranks
are balanced.

To demonstrate, we simulate a skewed version of the CKD data where normality
fails (e.g. pain scores, length of stay — common in clinical data).

```{r onesample-np-data}
set.seed(101)
# Simulate skewed outcome — e.g. urinary protein excretion (mg/day)
# Reference value in healthy adults: 150 mg/day
protein <- round(rexp(25, rate = 1/220), 0)  # right-skewed

cat("n      =", length(protein), "\n")
cat("Median =", median(protein), "mg/day\n")
cat("Mean   =", round(mean(protein), 1), "mg/day\n\n")

# Normality check — expect failure with skewed data
shapiro.test(protein)
```

```{r onesample-np-hist, fig.height=3.5}
ggplot(data.frame(protein), aes(x = protein)) +
  geom_histogram(bins = 12, fill = "#FAC775", colour = "#854F0B", alpha = 0.8) +
  geom_vline(xintercept = 150, linetype = "dashed", colour = "#D85A30",
             linewidth = 1) +
  annotate("text", x = 160, y = 4, label = "Reference\n150 mg/day",
           colour = "#D85A30", size = 3.5, hjust = 0) +
  labs(title   = "Urinary protein excretion in CKD patients (skewed)",
       x = "Protein excretion (mg/day)", y = "Count") +
  theme_minimal(base_size = 13)
```

```{r onesample-np-test}
# Wilcoxon signed-rank test — mu is the reference value (null median)
wilcox.test(protein, mu = 150, alternative = "two.sided", conf.int = TRUE)
```
V = 205.5, p-value = 0.2528, No significant difference. 

```{r onesample-np-effect}
# Effect size r = Z / sqrt(N) — common for Wilcoxon tests
wt <- wilcox.test(protein, mu = 150, alternative = "two.sided")
# Use rstatix for a clean effect size
wilcox_effsize_1s <- wilcox_effsize(
  data.frame(protein, ref = 150),
  protein ~ 1, mu = 150
)
cat("Effect size r =", round(wilcox_effsize_1s$effsize, 3), "\n")
cat("Interpretation: small < 0.1, medium < 0.3, large >= 0.5\n")
```

> **When to choose:** If Shapiro-Wilk p < 0.05 or the data are clearly skewed
> (histograms, boxplots), use `wilcox.test()`. Report the **median and IQR**
> rather than mean and SD. The Wilcoxon test is nearly as powerful as the
> t-test when normality holds, and more valid when it does not.

---

# Independent samples t-test

## Epidemiological scenario

A community health survey compares fasting blood glucose (FBG) between two
groups: adults with a **sedentary lifestyle** vs. those who are **physically
active** (≥ 150 min/week moderate activity). The outcome is FBG in mg/dL.

**Hypotheses:**
- H₀: μ_sedentary = μ_active
- H₁: μ_sedentary ≠ μ_active

```{r indep-data}

#simulate the data for each group
set.seed(7)
sedentary <- round(rnorm(35, mean = 102, sd = 15), 1)
active    <- round(rnorm(40, mean = 93,  sd = 13), 1)

df_fbg <- data.frame(
  fbg   = c(sedentary, active),
  group = factor(c(rep("Sedentary", 35), rep("Active", 40)),
                 levels = c("Sedentary", "Active"))
)

df_fbg %>%
  group_by(group) %>%
  summarise(n = n(), mean = round(mean(fbg), 2),
            sd = round(sd(fbg), 2), se = round(sd(fbg)/sqrt(n()), 2))
```

## Assumption checks

### 1. Normality (per group)

```{r indep-normality, fig.height=3.5}
df_fbg %>%
  group_by(group) %>%
  summarise(W = shapiro.test(fbg)$statistic,
            p = shapiro.test(fbg)$p.value)

ggqqplot(df_fbg, x = "fbg", facet.by = "group",
         title = "Q-Q plots — fasting blood glucose by activity group")
```

### 2. Homogeneity of variance — Levene's test

```{r levene}
leveneTest(fbg ~ group, data = df_fbg)
```

> **Decision rule:**
> - Levene p > 0.05 → variances are equal → use standard `var.equal = TRUE`
> - Levene p ≤ 0.05 → variances differ → use Welch correction (`var.equal = FALSE`, the default in R)

## Visualise

```{r indep-plot, fig.height=4}
ggplot(df_fbg, aes(x = group, y = fbg, fill = group)) +
  geom_boxplot(alpha = 0.7, width = 0.5, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  scale_fill_manual(values = c("Sedentary" = "#F0997B", "Active" = "#5DCAA5")) +
  labs(title   = "Fasting blood glucose by physical activity level",
       x = NULL, y = "FBG (mg/dL)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## Run the test

```{r indep-test}
# var.equal = FALSE uses Welch correction (safer default)
result_indep <- t.test(fbg ~ group, data = df_fbg, var.equal = FALSE)
print(result_indep)
```

## Effect size

```{r indep-effect}
cohens_d_indep <- df_fbg %>% cohens_d(fbg ~ group)
print(cohens_d_indep)
```

> **Conclusion:** Sedentary adults had significantly higher FBG than active
> adults. Report as: mean difference = X mg/dL (95% CI: [L, U]),
> t(58.55) = 3.8086, p = 0.00034, Cohen's d = 0.89.

## Non-parametric analogue — Mann-Whitney U test

Also called the Wilcoxon rank-sum test. Use when normality is violated in
either group. It compares the **rank distributions** of two independent groups
rather than their means — equivalent to testing whether a randomly selected
observation from group 1 tends to be larger than one from group 2.

To demonstrate, we simulate a scenario where hospital length of stay (LOS)
is compared between patients admitted via the emergency department vs. elective
admission — a classically right-skewed outcome.

```{r mwu-data}
set.seed(33)
los_emergency <- round(rexp(40, rate = 1/6.5))   # median ~5 days, right-skewed
los_elective  <- round(rexp(38, rate = 1/3.8))   # median ~3 days

df_los <- data.frame(
  los    = c(los_emergency, los_elective),
  route  = factor(c(rep("Emergency", 40), rep("Elective", 38)),
                  levels = c("Emergency", "Elective"))
)

# Summary: median and IQR more appropriate for skewed data
df_los %>%
  group_by(route) %>%
  summarise(n      = n(),
            median = median(los),
            IQR    = IQR(los),
            mean   = round(mean(los), 1))
```

```{r mwu-normality}
# Confirm normality violation
df_los %>%
  group_by(route) %>%
  summarise(W = round(shapiro.test(los)$statistic, 4),
            p = round(shapiro.test(los)$p.value, 4))
```

```{r mwu-plot, fig.height=4}
ggplot(df_los, aes(x = route, y = los, fill = route)) +
  geom_boxplot(alpha = 0.7, width = 0.5, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  scale_fill_manual(values = c("Emergency" = "#F0997B",
                                "Elective"  = "#5DCAA5")) +
  labs(title   = "Hospital length of stay by admission route",
       subtitle = "Right-skewed outcome — Mann-Whitney U preferred",
       x = NULL, y = "Length of stay (days)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

```{r mwu-test}
# Mann-Whitney U / Wilcoxon rank-sum test
wilcox.test(los ~ route, data = df_los,
            alternative = "two.sided", conf.int = TRUE)
```

```{r mwu-effect}
# Effect size r for Mann-Whitney U
mwu_eff <- df_los %>% wilcox_effsize(los ~ route)
print(mwu_eff)
```

> **Reporting:** Emergency admissions had significantly longer hospital stays
> than elective admissions (median X vs. X days; W = X, p = X, r = X).
> Report medians and IQRs — not means — for skewed data.

---

# Paired t-test

## Clinical scenario

A clinical trial evaluates whether an **8-week dietary intervention** reduces
LDL cholesterol. The same 25 patients are measured at baseline and after
8 weeks. This is a paired design — each patient is their own control.

**Hypotheses:**
- H₀: μ_difference = 0 (no change in LDL)
- H₁: μ_difference ≠ 0

```{r paired-data}
set.seed(15)
n_paired    <- 25
ldl_before  <- round(rnorm(n_paired, mean = 148, sd = 22), 1)
ldl_after   <- round(ldl_before - rnorm(n_paired, mean = 18, sd = 12), 1)
differences <- ldl_after - ldl_before

df_ldl <- data.frame(
  patient = 1:n_paired,
  before  = ldl_before,
  after   = ldl_after,
  diff    = differences
)

cat("Mean LDL before:", round(mean(ldl_before), 2), "mg/dL\n")
cat("Mean LDL after: ", round(mean(ldl_after),  2), "mg/dL\n")
cat("Mean difference:", round(mean(differences), 2), "mg/dL\n")
cat("SD of differences:", round(sd(differences), 2), "mg/dL\n")
```

## Assumption check

For a paired t-test, normality of the **differences** is required — not of
the raw measurements.

```{r paired-normality, fig.height=3.5}
shapiro.test(differences)

ggqqplot(differences,
         title = "Q-Q plot — LDL differences (after − before)")
```

## Visualise — spaghetti plot

```{r paired-spaghetti, fig.height=4}
df_long <- df_ldl %>%
  select(patient, before, after) %>%
  pivot_longer(cols = c(before, after),
               names_to = "time", values_to = "ldl") %>%
  mutate(time = factor(time, levels = c("before", "after")))

ggplot(df_long, aes(x = time, y = ldl, group = patient)) +
  geom_line(alpha = 0.35, colour = "#378ADD") +
  geom_point(aes(colour = time), size = 2.5, alpha = 0.8) +
  stat_summary(aes(group = 1), fun = mean, geom = "line",
               colour = "#D85A30", linewidth = 1.5) +
  stat_summary(aes(group = 1), fun = mean, geom = "point",
               colour = "#D85A30", size = 4) +
  scale_colour_manual(values = c("before" = "#85B7EB", "after" = "#1D9E75")) +
  scale_x_discrete(labels = c("before" = "Baseline", "after" = "8 weeks")) +
  labs(title   = "LDL cholesterol before and after dietary intervention",
       subtitle = "Blue lines = individual patients · Red = group mean",
       x = NULL, y = "LDL (mg/dL)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## Run the test

```{r paired-test}
result_paired <- t.test(ldl_after, ldl_before,
                        paired = TRUE, alternative = "two.sided")
print(result_paired)
```

## Effect size

```{r paired-effect}
d_paired <- mean(differences) / sd(differences)
cat("Cohen's d (paired) =", round(d_paired, 3), "\n")
```

> **Conclusion:** The dietary intervention produced a mean LDL reduction of
> X mg/dL (95% CI: [L, U]), which was statistically significant (t(24) = X,
> p < 0.05, Cohen's d = X). This represents a clinically meaningful reduction.

## Non-parametric analogue — Wilcoxon signed-rank test (paired)

Use when the **differences** between paired measurements are not normally
distributed. This is common with pain scores, quality-of-life scales, or small
samples. The test ranks the absolute differences and checks whether positive
ranks outweigh negative ranks.

We simulate pain scores (0–10 NRS) before and after an analgesic intervention
in 20 patients — ordinal data with a bounded range, where normality of
differences is unlikely.

```{r paired-np-data}
set.seed(77)
n_pain     <- 20
pain_before <- sample(5:9, n_pain, replace = TRUE)
pain_after  <- pmax(0, pain_before - sample(0:5, n_pain, replace = TRUE))
pain_diff   <- pain_after - pain_before

df_pain <- data.frame(
  patient     = 1:n_pain,
  before      = pain_before,
  after       = pain_after,
  diff        = pain_diff
)

cat("Median pain before:", median(pain_before), "\n")
cat("Median pain after: ", median(pain_after),  "\n")
cat("Median difference: ", median(pain_diff),   "\n\n")

shapiro.test(pain_diff)
```

```{r paired-np-plot, fig.height=4}
df_pain_long <- df_pain %>%
  pivot_longer(cols = c(before, after),
               names_to = "time", values_to = "pain") %>%
  mutate(time = factor(time, levels = c("before", "after")))

ggplot(df_pain_long, aes(x = time, y = pain, group = patient)) +
  geom_line(alpha = 0.35, colour = "#378ADD") +
  geom_point(aes(colour = time), size = 2.5, alpha = 0.8) +
  stat_summary(aes(group = 1), fun = median, geom = "line",
               colour = "#D85A30", linewidth = 1.5) +
  stat_summary(aes(group = 1), fun = median, geom = "point",
               colour = "#D85A30", size = 4) +
  scale_colour_manual(values = c("before" = "#85B7EB", "after" = "#1D9E75")) +
  scale_x_discrete(labels = c("before" = "Pre-treatment",
                               "after"  = "Post-treatment")) +
  scale_y_continuous(breaks = 0:10, limits = c(0, 10)) +
  labs(title    = "Pain scores before and after analgesic intervention",
       subtitle = "Blue lines = individual patients · Red = group median",
       x = NULL, y = "Pain score (NRS 0–10)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

```{r paired-np-test}
wilcox.test(pain_after, pain_before,
            paired = TRUE, alternative = "two.sided", conf.int = TRUE)
```

```{r paired-np-effect}
pain_long2 <- data.frame(
  pain = c(pain_before, pain_after),
  time = factor(rep(c("before", "after"), each = n_pain),
                levels = c("before", "after")),
  id   = rep(1:n_pain, 2)
)
paired_eff <- pain_long2 %>% wilcox_effsize(pain ~ time, paired = TRUE)
print(paired_eff)
```

> **Reporting:** Pain scores were significantly reduced after treatment
> (median X vs. X; V = X, p = X, r = X). When reporting Wilcoxon paired
> results, use V (the test statistic), not W, and report medians with IQR.

---

# One-way ANOVA

## Epidemiological scenario

A population survey measures **BMI** across four dietary pattern groups:
Western, Mediterranean, Vegetarian, and Traditional. The research question
is whether mean BMI differs across groups.

**Hypotheses:**
- H₀: μ_Western = μ_Mediterranean = μ_Vegetarian = μ_Traditional
- H₁: At least one group mean differs

```{r anova-data}
set.seed(99)
diet_data <- data.frame(
  bmi  = c(round(rnorm(40, 28.5, 4.2), 1),
            round(rnorm(38, 25.1, 3.8), 1),
            round(rnorm(35, 24.3, 3.5), 1),
            round(rnorm(42, 26.8, 4.0), 1)),
  diet = factor(rep(c("Western", "Mediterranean",
                       "Vegetarian", "Traditional"), c(40, 38, 35, 42)))
)

diet_data %>%
  group_by(diet) %>%
  summarise(n    = n(),
            mean = round(mean(bmi), 2),
            sd   = round(sd(bmi),   2),
            se   = round(sd(bmi) / sqrt(n()), 2)) %>%
  arrange(desc(mean))
```

## Assumption checks

### 1. Normality per group

```{r anova-normality}
diet_data %>%
  group_by(diet) %>%
  summarise(W = round(shapiro.test(bmi)$statistic, 4),
            p = round(shapiro.test(bmi)$p.value, 4))
```

```{r anova-qq, fig.height=5}
ggqqplot(diet_data, x = "bmi", facet.by = "diet",
         title = "Q-Q plots — BMI by dietary pattern")
```

### 2. Homogeneity of variance

```{r anova-levene}
leveneTest(bmi ~ diet, data = diet_data)
```

### 3. Visualise distributions

```{r anova-plot, fig.height=4.5}
diet_order <- diet_data %>%
  group_by(diet) %>%
  summarise(m = mean(bmi)) %>%
  arrange(m) %>%
  pull(diet)

diet_data$diet <- factor(diet_data$diet, levels = diet_order)

ggplot(diet_data, aes(x = diet, y = bmi, fill = diet)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_boxplot(width = 0.15, alpha = 0.8, outlier.shape = 21) +
  scale_fill_manual(values = c(
    "Western"       = "#F0997B",
    "Traditional"   = "#FAC775",
    "Mediterranean" = "#5DCAA5",
    "Vegetarian"    = "#9FE1CB")) +
  labs(title   = "BMI distribution by dietary pattern",
       x = NULL, y = "BMI (kg/m²)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## Run one-way ANOVA

```{r anova-run}
model_anova <- aov(bmi ~ diet, data = diet_data)
summary(model_anova)
```

## Effect size — η² (eta-squared)

```{r anova-eta}
# Correct usage - pass the model, not a formula
eta_sq <- eta_squared(model_anova)
print(eta_sq)

# Manual calculation
ss_between <- summary(model_anova)[[1]]["diet",   "Sum Sq"]
ss_total   <- sum(summary(model_anova)[[1]][, "Sum Sq"])
eta2       <- ss_between / ss_total

cat("η² =", round(eta2, 3),
    "| Interpretation:",
    ifelse(eta2 < 0.01, "negligible",
    ifelse(eta2 < 0.06, "small",
    ifelse(eta2 < 0.14, "medium", "large"))), "\n")
```

## Post-hoc tests — Tukey HSD

Run only because ANOVA is significant. Tukey HSD controls the family-wise
error rate across all pairwise comparisons.

```{r tukey}
tukey_result <- TukeyHSD(model_anova)
print(tukey_result)
```

```{r tukey-plot, fig.height=4}
tukey_df <- as.data.frame(tukey_result$diet) %>%
  tibble::rownames_to_column("comparison") %>%
  mutate(significant = `p adj` < 0.05)

ggplot(tukey_df, aes(x = reorder(comparison, diff),
                     y = diff, colour = significant)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "gray60", linewidth = 0.7) +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.2, linewidth = 0.8) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("TRUE" = "#1D9E75", "FALSE" = "#888780"),
                      labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05")) +
  coord_flip() +
  labs(title   = "Tukey HSD — pairwise BMI differences by diet group",
       x = NULL, y = "Mean difference in BMI (kg/m²)",
       colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

> **Conclusion:** One-way ANOVA revealed a significant difference in mean BMI
> across dietary pattern groups (F(3, X) = X, p < 0.05, η² = X). Post-hoc
> Tukey HSD identified significantly higher BMI in the Western diet group
> compared to Mediterranean and Vegetarian groups (p < 0.05).

## Non-parametric analogue — Kruskal-Wallis test

Use when normality is violated in one or more groups, or when the outcome is
ordinal. It is the rank-based extension of the Mann-Whitney U test to 3+
groups. Like ANOVA it tests whether at least one group differs, but it
compares **rank distributions** rather than means.

We simulate patient-reported quality of life (QoL) scores (0–100) across
three cancer treatment arms — an outcome that is often bounded, skewed, and
unlikely to be normally distributed.

```{r kw-data}
set.seed(44)
qol_data <- data.frame(
  qol = c(round(rbeta(35, 2, 5) * 100),    # Surgery — lower QoL
           round(rbeta(33, 3, 4) * 100),    # Chemo — moderate
           round(rbeta(36, 4, 3) * 100)),   # Targeted therapy — higher
  arm = factor(rep(c("Surgery", "Chemotherapy", "Targeted therapy"),
                   c(35, 33, 36)))
)

qol_data %>%
  group_by(arm) %>%
  summarise(n      = n(),
            median = median(qol),
            IQR    = IQR(qol),
            mean   = round(mean(qol), 1))
```

```{r kw-normality}
qol_data %>%
  group_by(arm) %>%
  summarise(W = round(shapiro.test(qol)$statistic, 4),
            p = round(shapiro.test(qol)$p.value, 4))
```

```{r kw-plot, fig.height=4.5}
ggplot(qol_data, aes(x = reorder(arm, qol, FUN = median),
                     y = qol, fill = arm)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_boxplot(width = 0.15, alpha = 0.85, outlier.shape = 21) +
  scale_fill_manual(values = c(
    "Surgery"           = "#F0997B",
    "Chemotherapy"      = "#FAC775",
    "Targeted therapy"  = "#5DCAA5")) +
  labs(title   = "Quality of life scores by cancer treatment arm",
       subtitle = "Skewed distributions — Kruskal-Wallis preferred",
       x = NULL, y = "QoL score (0–100)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

```{r kw-test}
kruskal.test(qol ~ arm, data = qol_data)
```

```{r kw-effect}
# Effect size epsilon-squared (η²_H) for Kruskal-Wallis
kw_eff <- qol_data %>% kruskal_effsize(qol ~ arm)
print(kw_eff)
cat("\nInterpretation: small < 0.06, medium < 0.14, large >= 0.14\n")
```

## Post-hoc — Dunn test with Bonferroni correction

```{r dunn}
dunn_result <- dunnTest(qol ~ arm, data = qol_data, method = "bonferroni")
print(dunn_result)
```

```{r dunn-plot, fig.height=3.5}
dunn_df <- dunn_result$res %>%
  mutate(significant = P.adj < 0.05)

ggplot(dunn_df, aes(x = reorder(Comparison, Z),
                    y = Z, colour = significant)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "gray60", linewidth = 0.7) +
  geom_point(size = 4) +
  geom_segment(aes(xend = reorder(Comparison, Z), yend = 0),
               linewidth = 1) +
  scale_colour_manual(values = c("TRUE" = "#1D9E75", "FALSE" = "#888780"),
                      labels = c("TRUE" = "p_adj < 0.05",
                                 "FALSE" = "p_adj ≥ 0.05")) +
  coord_flip() +
  labs(title   = "Dunn test — pairwise QoL comparisons (Bonferroni)",
       x = NULL, y = "Z statistic",
       colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

> **Reporting:** QoL scores differed significantly across treatment arms
> (Kruskal-Wallis H(2) = X, p = X, ε² = X). Dunn post-hoc tests with
> Bonferroni correction showed targeted therapy had significantly higher
> QoL than surgery (p_adj = X). Report medians and IQRs throughout.

---

# Statistical significance vs. clinical importance

A common error is equating a small p-value with a large or important effect.
Large samples can detect trivial differences.

```{r sig-vs-clinical}
set.seed(55)

# Same 2 mmHg difference — tiny samples (underpowered) vs. huge sample
results <- lapply(c(20, 50, 100, 500, 2000), function(n) {
  g1 <- rnorm(n, mean = 130, sd = 15)
  g2 <- rnorm(n, mean = 132, sd = 15)   # only 2 mmHg difference
  tt <- t.test(g1, g2)
  d  <- (mean(g2) - mean(g1)) / sqrt((var(g1) + var(g2)) / 2)
  data.frame(n = n, p_value = round(tt$p.value, 4),
             cohens_d = round(d, 3),
             significant = tt$p.value < 0.05)
})

do.call(rbind, results)
```

> **Key lesson:** The 2 mmHg difference is clinically negligible but becomes
> statistically significant at large n. Always report effect sizes alongside
> p-values. A significant p-value answers "is the difference real?" — not
> "does the difference matter clinically?"

---


# Exercises

**Exercise 1 
A hospital reports that average patient wait time in the emergency department
is 45 minutes. An audit of 28 randomly selected visits finds a mean of 52
minutes (SD = 11 min). Is there evidence that actual wait times exceed the
reported value? Use α = 0.05, one-tailed.

```{r ex1, eval=FALSE}
# Your code here — remember to check normality first

```

**Exercise 2 
A study compares haemoglobin (Hb) levels in anaemic patients randomised to
oral iron supplementation vs. IV iron infusion after 12 weeks:

```{r ex2-data}
set.seed(3)
oral_iron <- round(rnorm(30, mean = 10.8, sd = 1.6), 1)
iv_iron   <- round(rnorm(30, mean = 11.9, sd = 1.4), 1)
```

Check assumptions, run the appropriate test, report the result with effect size.

```{r ex2, eval=FALSE}
# Your code here

```

**Exercise 3 
Blood pressure is measured in 20 hypertensive patients before and 6 weeks
after starting an ACE inhibitor:

```{r ex3-data}
set.seed(8)
bp_before <- round(rnorm(20, mean = 158, sd = 12), 1)
bp_after  <- round(bp_before - rnorm(20, mean = 14, sd = 9), 1)
```

Check that the differences are normally distributed, run a test,
and construct a spaghetti plot.

```{r ex3, eval=FALSE}
# Your code here

```

**Exercise 4 
A clinical trial randomises 120 patients with type 2 diabetes to three
treatments: metformin, lifestyle modification, or combination therapy.
HbA1c (%) is measured after 6 months:

```{r ex4-data}
set.seed(22)
hba1c_data <- data.frame(
  hba1c = c(round(rnorm(40, 7.4, 0.9), 1),
             round(rnorm(40, 7.1, 0.8), 1),
             round(rnorm(40, 6.6, 0.7), 1)),
  treatment = factor(rep(c("Metformin", "Lifestyle", "Combination"), each = 40))
)
```

Check assumptions, run test, report effect size, and identify which treatments differ. Produce a boxplot with significance annotations.

```{r ex4, eval=FALSE}
# Your code here

```

**Exercise 5 
A published paper reports: *"Mean systolic BP was 2.1 mmHg lower in the
treatment group (p = 0.003)."* The study enrolled n = 8,000 participants.

- What is the likely explanation for the highly significant p-value?
- Calculate Cohen's d assuming SD = 15 mmHg. Interpret it.
- What additional information would you want before judging clinical importance?

```{r ex5, eval=FALSE}
# Calculate Cohen's d and reflect in comments

```

---

**Exercise 6 
A study compares time-to-discharge (days) in patients receiving standard care
vs. an accelerated rehabilitation programme after hip replacement surgery:

```{r ex6-data}
set.seed(66)
standard_care <- round(rexp(28, rate = 1/7.5))
accel_rehab   <- round(rexp(28, rate = 1/5.2))
```

Confirm normality is violated, run the appropriate test, calculate the
effect size, and produce a boxplot with the median labelled.

```{r ex6, eval=FALSE}
# Your code here

```

**Exercise 7 
Anxiety scores (GAD-7, range 0–21) are recorded in 18 patients before and
after an 6-week mindfulness programme:

```{r ex7-data}
set.seed(88)
gad_before <- sample(10:18, 18, replace = TRUE)
gad_after  <- pmax(0, gad_before - sample(0:8, 18, replace = TRUE))
```

Check normality of differences. Run the appropriate test and produce a
spaghetti plot with the median trajectory highlighted.

```{r ex7, eval=FALSE}
# Your code here

```

**Exercise 8 
Serum ferritin levels (μg/L) are compared across four iron deficiency severity
groups: none, mild, moderate, severe:

```{r ex8-data}
set.seed(11)
ferritin_data <- data.frame(
  ferritin  = c(round(rexp(25, rate = 1/120)),
                round(rexp(25, rate = 1/55)),
                round(rexp(25, rate = 1/22)),
                round(rexp(25, rate = 1/8))),
  severity  = factor(rep(c("None", "Mild", "Moderate", "Severe"), each = 25),
                     levels = c("None", "Mild", "Moderate", "Severe"))
)
```

Confirm normality, run test, calculate effect size, and identify which
severity groups differ.

```{r ex8, eval=FALSE}
# Your code here

```


---

# Session info

```{r session}
sessionInfo()
```
