---
title: "Sample Size Calculation — Comparing Means"
subtitle: Fundamentals of Biostatistics
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

---

# Setup

Install and load the required packages. The `pwr` package is the standard R tool
for power and sample size calculations. We also use `ggplot2` for visualizations.

```{r packages}
# Install if not already installed
if (!require(pwr))     install.packages("pwr")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(dplyr))   install.packages("dplyr")

library(pwr)
library(ggplot2)
library(dplyr)
```

---

# Key concepts recap

Before calculating, confirm you can specify all four inputs:

| Input            | Symbol       | Meaning                                         |
|------------------|--------------|-------------------------------------------------|
| Significance level | α          | False positive rate (typically 0.05)            |
| Power            | 1 − β        | Probability of detecting a real effect (≥ 0.80) |
| Effect size      | d = Δ / σ    | Minimum clinically important difference / SD    |
| Sample size      | n            | Per group (for two-sample tests)                |

> **Key insight:** Give any three of these four, and you can solve for the fourth.
> In practice we solve for **n**, but you can also solve for **power** given a
> fixed n (useful when budget constrains your sample size).

---

# One-sample t-test

**Scenario:** You want to test whether the mean systolic blood pressure (SBP) in
a patient group differs from a known population mean of 120 mmHg. From pilot
data, σ ≈ 15 mmHg. You consider a 5 mmHg difference clinically meaningful.

## Calculate sample size

```{r one-sample}
# Inputs
mu_0      <- 120    # hypothesised population mean (mmHg)
mu_1      <- 125    # expected mean in study group
sigma     <- 15     # standard deviation
alpha     <- 0.05   # significance level (two-tailed)
power     <- 0.80   # desired power

# Cohen's d for one-sample
delta <- mu_1 - mu_0
d     <- delta / sigma

cat("Minimum clinically important difference (Δ):", delta, "mmHg\n")
cat("Standard deviation (σ):", sigma, "mmHg\n")
cat("Cohen's d:", round(d, 3), "\n\n")

# pwr.t.test: type = "one.sample"
result_1s <- pwr.t.test(
  d    = d,
  sig.level = alpha,
  power     = power,
  type      = "one.sample",
  alternative = "two.sided"
)

print(result_1s)
```

```{r one-sample-ceiling}
# Always round UP to the nearest whole number
n_required <- ceiling(result_1s$n)
cat("Required sample size:", n_required, "participants\n")
```

---

# Two-sample independent t-test

**Scenario:** An RCT compares mean SBP between a treatment group and a control
group. You expect the treatment to reduce SBP by 8 mmHg. Both groups are assumed
to have σ = 14 mmHg.

## Calculate sample size

```{r two-sample}
# Inputs
delta  <- 8     # minimum clinically important difference (mmHg)
sigma  <- 14    # common SD in both groups
alpha  <- 0.05
power  <- 0.80

d <- delta / sigma
cat("Cohen's d:", round(d, 3), "\n\n")

# pwr.t.test: type = "two.sample"
result_2s <- pwr.t.test(
  d           = d,
  sig.level   = alpha,
  power       = power,
  type        = "two.sample",
  alternative = "two.sided"
)

print(result_2s)
cat("\nRequired per group:", ceiling(result_2s$n), "\n")
cat("Total sample size: ", ceiling(result_2s$n) * 2, "\n")
```

## Manual formula verification

The formula:
$$n = \frac{2\sigma^2(z_{\alpha/2} + z_\beta)^2}{\Delta^2}$$

```{r manual-formula}
z_alpha <- qnorm(1 - alpha / 2)   # 1.96 for alpha = 0.05
z_beta  <- qnorm(power)            # 0.84 for power = 0.80

n_manual <- (2 * sigma^2 * (z_alpha + z_beta)^2) / delta^2

cat("z_alpha/2 =", round(z_alpha, 3), "\n")
cat("z_beta    =", round(z_beta,  3), "\n")
cat("n per group (formula):", ceiling(n_manual), "\n")
cat("n per group (pwr):    ", ceiling(result_2s$n), "\n")
```

> The small difference between the manual formula and `pwr.t.test` arises because
> `pwr` uses the non-central t-distribution for greater accuracy. The formula is
> a good approximation; `pwr` is preferred in practice.

---

# Power curve — given a fixed sample size

**Scenario:** Your budget only allows n = 50 per group. What power do you
achieve across a range of effect sizes?

```{r power-curve}
n_fixed <- 50

power_curve <- data.frame(
  d = seq(0.1, 1.0, by = 0.01)
) %>%
  rowwise() %>%
  mutate(
    power = pwr.t.test(
      n = n_fixed, d = d,
      sig.level = 0.05,
      type = "two.sample",
      alternative = "two.sided"
    )$power
  ) %>%
  ungroup()

ggplot(power_curve, aes(x = d, y = power)) +
  geom_line(colour = "#378ADD", linewidth = 1.2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "#D85A30") +
  geom_hline(yintercept = 0.90, linetype = "dashed", colour = "#BA7517") +
  annotate("text", x = 0.12, y = 0.82, label = "80% power", size = 3.5,
           colour = "#D85A30", hjust = 0) +
  annotate("text", x = 0.12, y = 0.92, label = "90% power", size = 3.5,
           colour = "#BA7517", hjust = 0) +
  labs(
    title    = paste("Power curve — n =", n_fixed, "per group"),
    subtitle = "Two-sample t-test · α = 0.05",
    x        = "Cohen's d",
    y        = "Power (1 − β)"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal(base_size = 13)
```

```{r power-at-d}
# What effect size achieves 80% power with n = 50?
min_d <- pwr.t.test(
  n = n_fixed, power = 0.80,
  sig.level = 0.05,
  type = "two.sample",
  alternative = "two.sided"
)$d

cat("Minimum detectable Cohen's d with n =", n_fixed, "per group:", round(min_d, 3), "\n")
cat("In mmHg (assuming σ = 14):", round(min_d * 14, 1), "mmHg\n")
```

---

# The power–sample size trade-off

How does increasing power from 80% to 90% affect n?

```{r power-tradeoff}
powers <- c(0.70, 0.80, 0.85, 0.90, 0.95)
delta  <- 8
sigma  <- 14
d      <- delta / sigma

tradeoff <- data.frame(power = powers) %>%
  rowwise() %>%
  mutate(
    n_per_group = ceiling(pwr.t.test(
      d = d, sig.level = 0.05, power = power,
      type = "two.sample", alternative = "two.sided"
    )$n),
    total_n = n_per_group * 2
  ) %>%
  ungroup() %>%
  mutate(
    pct_increase = round((n_per_group / n_per_group[power == 0.80] - 1) * 100, 0)
  )

print(tradeoff)
```

```{r power-tradeoff-plot}
ggplot(tradeoff, aes(x = factor(paste0(power * 100, "%")), y = n_per_group)) +
  geom_col(fill = "#378ADD", alpha = 0.85, width = 0.6) +
  geom_text(aes(label = n_per_group), vjust = -0.5, size = 4) +
  labs(
    title    = "Sample size per group by target power",
    subtitle = "Δ = 8 mmHg · σ = 14 mmHg · α = 0.05",
    x        = "Target power",
    y        = "n per group"
  ) +
  theme_minimal(base_size = 13)
```

---

# Practical adjustments

## Adjust for expected dropout

```{r dropout}
n_base      <- ceiling(result_2s$n)
dropout_rate <- 0.15   # 15% expected dropout

n_adjusted <- ceiling(n_base / (1 - dropout_rate))

cat("Base n per group:     ", n_base, "\n")
cat("Dropout rate:         ", dropout_rate * 100, "%\n")
cat("Adjusted n per group: ", n_adjusted, "\n")
cat("Adjusted total n:     ", n_adjusted * 2, "\n")
```

## Adjust for cluster design

```{r cluster}
n_simple <- ceiling(result_2s$n)
deff     <- 1.8   # design effect — typical for community cluster trials

n_cluster <- ceiling(n_simple * deff)

cat("Simple random n per group:", n_simple, "\n")
cat("Design effect (DEFF):     ", deff, "\n")
cat("Cluster-adjusted n:       ", n_cluster, "\n")
```

---

# Exercises

Work through the following. Use `pwr.t.test()` for each.

**Exercise 1.**
A study compares mean fasting glucose between diabetic patients on two different
treatment regimens. Assume σ = 20 mg/dL and a clinically meaningful difference
of 10 mg/dL. Calculate n per group for α = 0.05 and 80% power.

```{r ex1, eval=FALSE}
# Your code here

```

**Exercise 2.**
Using the same scenario as Exercise 1, calculate the power achieved if the study
is limited to n = 60 per group due to budget constraints.

```{r ex2, eval=FALSE}
# Your code here

```

**Exercise 3.**
Plot a sensitivity analysis showing how n per group changes as σ varies from
15 to 30 mg/dL, keeping Δ = 10 mg/dL, α = 0.05, and power = 80%.

```{r ex3, eval=FALSE}
# Your code here

```

**Exercise 4.**
The study in Exercise 1 expects 20% dropout and will use cluster sampling
with DEFF = 2.0. What is the final recommended total sample size?

```{r ex4, eval=FALSE}
# Your code here

```

---

# Session info

```{r session}
sessionInfo()
```
