---
title: "Introduction to Survival Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width  = 3,
  fig.height = 5
)
```

---

# Overview

This notebook introduces the core methods of **survival analysis** (time-to-event analysis) using real medical datasets from the `survival` package. Topics covered:

| Section | Method | Dataset | Clinical context |
|---|---|---|---|
| 2 | Survival objects & censoring | `lung` | Lung cancer mortality |
| 3 | Kaplan-Meier estimation | `lung` | Survival by sex |
| 4 | Log-rank test | `pbc` | Liver disease RCT |
| 5 | Cox proportional hazards | `lung` | Multivariable adjustment |
| 6 | PH assumption diagnostics | `lung` | Schoenfeld residuals |


---

# Setup & Packages

```{r packages}
if (!requireNamespace("survival",  quietly = TRUE)) install.packages("survival")
if (!requireNamespace("survminer", quietly = TRUE)) install.packages("survminer")
if (!requireNamespace("ggplot2",   quietly = TRUE)) install.packages("ggplot2")
if (!requireNamespace("dplyr",     quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("knitr",     quietly = TRUE)) install.packages("knitr")
if (!requireNamespace("broom",     quietly = TRUE)) install.packages("broom")
if (!requireNamespace("gridExtra", quietly = TRUE)) install.packages("gridExtra")

library(survival)   # Surv(), survfit(), coxph(), survreg(), cox.zph()
library(survminer)  # ggsurvplot() — publication-ready KM plots
library(ggplot2)    # additional visualizations
library(dplyr)      # data manipulation
library(knitr)      # kable() for formatted tables
library(broom)      # tidy() for model outputs
library(gridExtra)  # multi-panel plots

theme_set(theme_minimal(base_size = 10))
```

---

# Log-Rank Test

## The `pbc` Dataset — Primary Biliary Cholangitis RCT

The `pbc` dataset records a **double-blind randomised trial** of D-penicillamine vs. placebo in 418 patients with **primary biliary cholangitis** (formerly cirrhosis) at the Mayo Clinic (1974–1984). Status 2 = transplant-free death.

```{r pbc-load}
data(pbc, package = "survival")

# Clean dataset — restrict to randomised patients (non-NA trt)
pbc2 <- pbc |>
  filter(!is.na(trt)) |>
  mutate(
    status2 = ifelse(status == 2, 1, 0),    # 1 = dead, 0 = censored/transplant
    trt     = factor(trt, levels = c(1, 2),
                     labels = c("D-penicillamine", "Placebo")),
    stage   = factor(stage),
    sex     = factor(sex, levels = c("m", "f"),
                     labels = c("Male", "Female")),
    bili_cat = cut(bili, breaks = c(0, 1, 3, 10, Inf),
                   labels = c("≤1", "1–3", "3–10", ">10"),
                   right = TRUE)
  )

cat("Randomised patients:", nrow(pbc2), "\n")
cat("Events (deaths):    ", sum(pbc2$status2), "\n")
cat("Censored:           ", sum(pbc2$status2 == 0), "\n")
```

## KM Curves by Treatment

```{r pbc-km}
km_pbc <- survfit(Surv(time, status2) ~ trt, data = pbc2)

ggsurvplot(
  km_pbc,
  data             = pbc2,
  conf.int         = TRUE,
  risk.table       = TRUE,
  risk.table.height = 0.25,
  pval             = TRUE,
  pval.method      = TRUE,
  xlab             = "Time (days)",
  ylab             = "Transplant-free survival probability",
  title            = "Survival by Treatment — PBC Trial (Mayo Clinic)",
  legend.labs      = c("D-penicillamine", "Placebo"),
  palette          = c("#d7301f", "#2c7fb8"),
  ggtheme          = theme_minimal(base_size = 12),
  surv.median.line  = "hv",
  tables.theme      = theme_cleantable(),
  fontsize          = 3,
  font.title        = c(11, "bold"),
  font.subtitle     = c(9, "plain"),
  font.x            = c(9, "plain"),
  font.y            = c(9, "plain"),
  font.tickslab     = c(7, "plain"),
  font.legend       = c(8, "plain"),
  pval.size         = 3.5,
  pval.coord        = c(10, 0.15)
)
```

## Formal Log-Rank Test

```{r logrank-pbc}
lr_pbc <- survdiff(Surv(time, status2) ~ trt, data = pbc2)
print(lr_pbc)

# Extract key values
chi2 <- lr_pbc$chisq
df   <- length(lr_pbc$n) - 1
pval <- pchisq(chi2, df = df, lower.tail = FALSE)

cat(sprintf("\nLog-rank test: χ²(%d) = %.3f,  p = %.4f\n", df, chi2, pval))
cat(sprintf("Interpretation: %s\n",
    ifelse(pval < 0.05,
           "Reject H0 — the two survival curves differ significantly.",
           "Fail to reject H0 — no significant difference between curves.")))
```
Log-rank test: χ²(1) = 0.102,  p = 0.7498
Interpretation: Fail to reject H0 — no significant difference between curves.


## Log-Rank by Bilirubin Category

```{r logrank-bili}
km_bili <- survfit(Surv(time, status2) ~ bili_cat, data = pbc2)
lr_bili <- survdiff(Surv(time, status2) ~ bili_cat, data = pbc2)

cat(sprintf("Log-rank (bilirubin categories): χ²(%d) = %.3f,  p < 0.001\n",
            length(lr_bili$n) - 1, lr_bili$chisq))

ggsurvplot(
  km_bili,
  data             = pbc2,
  conf.int         = FALSE,
  risk.table       = TRUE,
  risk.table.height = 0.28,
  pval             = TRUE,
  xlab             = "Time (days)",
  ylab             = "Survival probability",
  title            = "Survival by Serum Bilirubin Level (PBC)",
  subtitle         = "Higher bilirubin strongly predicts worse survival",
  legend.labs      = c("≤1 mg/dL","1–3 mg/dL","3–10 mg/dL",">10 mg/dL"),
  palette          = c("#41ab5d","#2c7fb8","#fd8d3c","#d7301f"),
  ggtheme          = theme_minimal(base_size = 12),
  tables.theme      = theme_cleantable(),
  fontsize          = 3,
  font.title        = c(11, "bold"),
  font.subtitle     = c(9, "plain"),
  font.x            = c(9, "plain"),
  font.y            = c(9, "plain"),
  font.tickslab     = c(7, "plain"),
  font.legend       = c(8, "plain"),
  pval.size         = 3.5,
  pval.coord        = c(10, 0.15)
)
```

---

# Cox Proportional Hazards Model


## Multivariable Cox Model

```{r cox-multi}
cox_multi <- coxph(
  Surv(time, status2) ~ trt + sex + age + bili+ albumin + stage,
  data = pbc2
)
summary(cox_multi)
```

```{r cox-hr-table}
# Clean HR table with CIs
cox_tidy <- tidy(cox_multi, exponentiate = TRUE, conf.int = TRUE)

cox_tidy
```

```{r cox-forest}
# Forest plot
cox_plot_df <- cox_tidy |>
  mutate(
    term = dplyr::recode(term,
      "sexFemale"  = "Sex: Female vs. Male",
      "age"        = "Age (per year)",
      "ph.ecog"    = "ECOG score (per unit)",
      "ph.karno"   = "Karnofsky score: physician (per unit)",
      "wt.loss"    = "Weight loss (per lb)"
    )
  )

ggplot(cox_plot_df, aes(x = estimate, y = reorder(term, estimate))) +
  geom_vline(xintercept = 1, linetype = "dashed",
             color = "gray50", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.25, color = "#2c7fb8", linewidth = 0.9) +
  geom_point(aes(color = p.value < 0.05), size = 3.5) +
  scale_color_manual(
    values = c("TRUE" = "#d7301f", "FALSE" = "#2c7fb8"),
    labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05"),
    name = NULL
  ) +
  scale_x_log10(breaks = c(0.3, 0.5, 1.0, 1.5, 2.0, 3.0)) +
  labs(
    title    = "Hazard Ratios: Multivariable Cox Model (lung dataset)",
    subtitle = "HR < 1 = protective; HR > 1 = increased hazard. Error bars = 95% CI",
    x = "Hazard Ratio (log scale)", y = NULL
  ) +
  theme(legend.position = "top")
```
#Interpretation & Documentation


---

# Proportional Hazards Assumption: Testing & Diagnostics

## Why This Matters

The Cox model assumes that hazard ratios are **constant over time**. If a treatment is beneficial early but harmful late (or vice versa), the PH assumption is violated and a single HR is misleading.

## Method 1 — Log-Log Plot

```{r ph-loglog}
# Plot ln(-ln(S(t))) vs ln(t) — parallel lines = PH holds
km_sex_plot <- survfit(Surv(time, status2) ~ sex, data = pbc2)

# Extract strata sizes safely
strata_sizes <- km_sex_plot$strata  # named integer vector, e.g. sex=Male: 113, sex=Female: 54

# Check what labels actually exist before plotting
ll_df <- data.frame(
  time = km_sex_plot$time,
  surv = km_sex_plot$surv,
  sex  = rep(names(strata_sizes), times = strata_sizes)  # ✅ no diff() needed
) |>
  filter(surv > 0 & surv < 1) |>
  mutate(
    log_time         = log(time),
    log_neg_log_surv = log(-log(surv))
  )

ggplot(ll_df, aes(x = log_time, y = log_neg_log_surv, color = sex)) +
  geom_step(linewidth = 1) +
  scale_color_manual(values = c("sex=Male" = "#d7301f", "sex=Female" = "#2c7fb8")) +
  labs(
    title    = "Log-Log Plot: Testing Proportional Hazards Assumption",
    subtitle = "Parallel lines = PH holds | Crossing/diverging lines = PH violated",
    x        = "log(Time)", y = "log(−log(Ŝ(t)))", color = "Sex"
  ) +
  theme_minimal(base_size = 8) +
  theme(legend.position = "top")
```



## Method 2 — Schoenfeld Residuals

```{r ph-schoenfeld}
# cox.zph() tests PH via Schoenfeld residuals
cox_ph_test <- cox.zph(cox_multi)

cat("Schoenfeld Residuals Test for Proportional Hazards:\n")
print(cox_ph_test)

cat("\nInterpretation:\n")
cat("  A significant p-value (< 0.05) for a predictor indicates\n")
cat("  that its effect is NOT constant over time → PH violated.\n")
```
Violated for ph.karno only,  p= 0.010. 

```{r ph-schoenfeld-plots}
# Plot Schoenfeld residuals for significant predictors
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

plot(cox_ph_test[3],
     col  = "#2c7fb8", lwd = 1,
     xlab = "Time", ylab = "Beta(t) for bili")
abline(h = coef(cox_multi)["bili"],
       lty = 1, col = "gray40")

plot(cox_ph_test[5],
     col  = "#2c7fb8", lwd = 1,
     xlab = "Time", ylab = "stage")
abline(h = coef(cox_multi)["stage"],
       lty = 1, col = "gray40")

par(mfrow = c(1, 1))
```


## Method 3 — Observed vs. Predicted Survival

```{r ph-observed-predicted}
# Compare KM (observed) with Cox-predicted survival
cox_base <- survfit(cox_multi)

km_obs <- survfit(Surv(time, status2) ~ 1, data = pbc2)

# Build comparison dataframe
obs_df  <- data.frame(time = km_obs$time,
                      surv = km_obs$surv,
                      type = "Observed (KM)")
pred_df <- data.frame(time = cox_base$time,
                      surv = cox_base$surv,
                      type = "Predicted (Cox)")

ggplot(rbind(obs_df, pred_df), aes(x = time, y = surv, color = type)) +
  geom_step(linewidth = 1) +
  scale_color_manual(
    values = c("Observed (KM)" = "#2c7fb8",
               "Predicted (Cox)" = "#d7301f")
  ) +
  labs(
    title    = "Observed (KM) vs. Predicted (Cox) Survival",
    subtitle = "Close agreement indicates adequate model fit",
    x = "Time (days)", y = "Survival probability", color = NULL
  ) +
  theme(legend.position = "top")
```



```{r cox-influential}
# Influential observations via dfbeta
dfbeta_res <- residuals(cox_complete, type = "dfbeta")

dfbeta_df <- data.frame(
  id    = seq_len(nrow(lung2_complete)),
  sex   = dfbeta_res[, 1],
  age   = dfbeta_res[, 2],
  ecog  = dfbeta_res[, 3]
)

ggplot(dfbeta_df, aes(x = id, y = sex)) +
  geom_col(fill = "#2c7fb8", alpha = 0.7) +
  geom_hline(yintercept = c(-0.05, 0.05),
             linetype = "dashed", color = "#d7301f") +
  labs(
    title    = "dfbeta Residuals for Sex Coefficient",
    subtitle = "Large |dfbeta| → influential observation for that coefficient",
    x = "Patient index", y = "dfbeta (sex)"
  )
```

---


## Key Decisions Flowchart

```{r decision-flow}
cat("
  START
    │
    ▼
  Is the outcome time-to-event with censoring?
    └─ Yes → Survival analysis is appropriate
    └─ No  → Consider logistic (binary) or Poisson (count)
    │
    ▼
  STEP 1: Always plot the Kaplan-Meier curve first
    └─ Use ggsurvplot() with risk table
    │
    ▼
  STEP 2: Comparing 2+ groups?
    └─ Log-rank test (survdiff())
    └─ Report χ², p-value, and KM curves together
    │
    ▼
  STEP 3: Need to adjust for confounders?
    └─ Cox proportional hazards model (coxph())
    └─ Report HRs with 95% CI
    │
    ▼
  STEP 4: Always test the PH assumption
    └─ cox.zph() — Schoenfeld residuals
    └─ Log-log plot
    └─ If violated → stratify, time-varying coefficients,
                     or parametric model
    │
    ▼
  STEP 5: Multiple event types?
    └─ Use competing risks framework
    └─ Do NOT treat competing events as censored
    │
    ▼
  REPORT: KM plot + risk table | HRs with CI | PH test result
          Median survival with 95% CI | Concordance (C-statistic)
")
```

---

# Session Information

```{r session-info}
sessionInfo()
```
